---
title: "micromort: Curated Micromort and Microlife Risk Datasets"
format: gfm
---

<!-- README.md is generated from README.qmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
devtools::load_all()
library(ggplot2)
library(dplyr)
```

# micromort

A **data package** providing curated datasets of micromort (acute risk) and microlife (chronic risk) values from authoritative sources including Wikipedia, CDC MMWR, and academic literature.
## Features

- **62 acute risks** measured in micromorts (one-in-a-million death probability per event)
- **35+ chronic factors** measured in microlives (30-minute life expectancy change per day)
- **14 authoritative sources** with full provenance tracking
- **Parquet datasets** for cross-language compatibility (R, Python, Arrow)
- **Plumber REST API** for programmatic access
- **Interactive dashboard** for data exploration
- **targets pipeline** for reproducible data updates

## Installation

### R-Universe (Recommended)

```r
# Install from r-universe (pre-built binaries, fast)
install.packages("micromort", repos = "https://johngavin.r-universe.dev")
```

### GitHub

```r
# Install from GitHub (source)
# install.packages("devtools")
devtools::install_github("JohnGavin/micromort")
```

### Nix Users

This project provides a reproducible Nix environment with pre-built binaries from the `johngavin` cachix cache:

```bash
# Enter project shell (uses cachix cache for fast builds)
./default.sh

# Start R
R
```

#### Using with rix

To include micromort in your own Nix-based R project:

```r
library(rix)

rix(
  r_pkgs = c("dplyr", "ggplot2"),
  git_pkgs = list(
    list(
      package_name = "micromort",
      repo_url = "https://github.com/JohnGavin/micromort",
      commit = "main"  # Or specific SHA for reproducibility
    )
  ),
  project_path = ".",
  overwrite = TRUE
)
```

## Concepts

### Micromort (Acute Risk)
A **micromort** is a unit of mortality risk equal to a **one-in-a-million probability of death per specific event**.

- **Unit:** 1 micromort = 1/1,000,000 = 0.0001% death probability
- **Scope:** Per discrete event (e.g., one skydive, one surgery, one flight)
- **Sign:** Always non-negative (it's a probability)
- **Example:** Skydiving has ~8 micromorts per jump (8-in-a-million death chance per jump)

### Microlife (Chronic Risk)

A **microlife** is a unit of life expectancy change equal to **30 minutes of expected lifespan, measured per day** of exposure.

- **Unit:** 1 microlife = 30 minutes of life expectancy
- **Scope:** Per day of maintaining a habit or exposure
- **Sign:** Positive (life gained) or negative (life lost)
- **Example:** Smoking 2 cigarettes daily costs -1 microlife/day (losing 30 mins of life expectancy each day you smoke)

### Conversion {#conversion}

**1 micromort ≈ 0.7 microlives** (assuming 40 years remaining life expectancy).
The conversion scales linearly with remaining life expectancy:

| Remaining life expectancy | 1 micromort ≈ |
|---------------------------|---------------|
| 10 years | 0.18 microlives |
| 20 years | 0.35 microlives |
| 40 years (default) | 0.70 microlives |
| 60 years | 1.05 microlives |

Use `lle(prob, life_expectancy = ...)` and `as_microlife()` to convert at any age.
See the [Age-Based Hazard Rates](#calculate-baseline-risk-by-age) section for daily micromort exposure by age.

### Morbidity Metrics

Micromorts and microlives focus on mortality. For quality-of-life impacts:

- **QALY (Quality-Adjusted Life Year):** 1 year of perfect health. Used in healthcare economics.
- **DALY (Disability-Adjusted Life Years):** Disease burden combining:
    - **YLL (Years of Life Lost):** Premature mortality component
    - **YLD (Years Lived with Disability):** Morbidity component
- **QALD (Quality-Adjusted Life Days):** 1 day of perfect health. For common illnesses.

See the [Introduction vignette](https://johngavin.github.io/micromort/articles/introduction.html) for detailed examples and the [Glossary](#glossary) for all acronym definitions.

## Quick Start

### Load the Datasets

```{r load-data}
library(micromort)

# Load acute risks (micromorts per event)
acute <- load_acute_risks()
nrow(acute)

# Load chronic risks (microlives per day)
chronic <- load_chronic_risks()
nrow(chronic)

# Load source registry
sources <- load_sources()
nrow(sources)
```

### Acute Risks (Micromorts per Event)

```{r acute}
# Convert a probability to micromorts
# Example: 1 in 10,000 chance = 100 micromorts
as_micromort(1/10000)

# Top 10 riskiest activities (micromorts per event/period)
acute |>
  dplyr::select(activity, micromorts, category, period) |>
  head(10)
```

**Caption:** Micromorts represent probability of death per event. Higher values = higher risk per occurrence.

### Chronic Risks (Microlives per Day)

```{r chronic}
# Factors that reduce life expectancy (microlives lost per day)
chronic |>
  dplyr::filter(direction == "loss") |>
  dplyr::select(factor, microlives_per_day, category) |>
  head(10)
```

**Caption:** Microlives per day. Negative values = life expectancy loss; positive = gain. Effects accumulate daily.

### Visualize Risks

```{r plot, fig.cap="Risk comparison in micromorts (log scale). Bars show death probability per event. COVID-19 and other risks shown in separate panels for clarity.", fig.height=10}
# Filter to show only activities with micromorts >= 1 for clarity on log scale
plot_risks(common_risks() |> dplyr::filter(micromorts >= 1))
```

## Analysis Functions

### Compare Lifestyle Interventions

```{r compare}
# Compare quitting smoking vs losing weight (microlives gained per day)
compare_interventions(list(
  "Quit 10 cigarettes/day" = list(factor = "Smoking 10 cigarettes", change = -1),
  "Lose 5kg" = list(factor = "Being 5 kg overweight", change = -1)
))
```

### Calculate Baseline Risk by Age

```{r hazard}
# Daily baseline mortality risk at age 35 (micromorts per day just from being alive)
daily_hazard_rate(35)
```

### Lifestyle Tradeoffs

```{r tradeoff}
# How much exercise offsets smoking? (in microlives per day)
lifestyle_tradeoff("Smoking 2 cigarettes", "20 min moderate exercise")
```

## API Access

Launch the REST API for programmatic access:

```r
launch_api()
# API available at http://localhost:8080
# Swagger docs at http://localhost:8080/__docs__/
```

**Core endpoints** (30 total — see [REST API vignette](https://johngavin.github.io/micromort/articles/rest_api.html) for full reference):

- `GET /v1/risks/acute` — Acute risks (micromorts per event)
- `GET /v1/risks/chronic` — Chronic risks (microlives per day)
- `GET /v1/risks/cancer` — Cancer mortality by type/sex/age
- `GET /v1/analysis/equivalence` — Risk equivalence lookup
- `GET /v1/convert/hazard-rate?age=35` — Daily hazard rate
- `GET /v1/sources` — Source registry
- `GET /health` — Health check

## Interactive Dashboard

Launch the Shiny dashboard:

```r
launch_dashboard()
```

## Risk Quiz

[Play in browser](https://johngavin.github.io/micromort/articles/quiz_shinylive.html)
— runs via WebR/Shinylive (30-60s initial load). Or locally:

```r
micromort::launch_quiz()
```

## Data Sources

| Source | Type | Data |
|--------|------|------|
| Wikipedia: Micromort | Encyclopedia | ~50 acute risks |
| Wikipedia: Microlife | Encyclopedia | ~20 chronic risks |
| micromorts.rip | Database | ~45 acute risks |
| CDC MMWR | Government | COVID vaccine data |
| Spiegelhalter (2012) BMJ | Academic | Microlife framework |
| SEER Cancer Statistics | Government | Cancer mortality by type/sex |

## Project Structure

<details>
<summary>Click to expand project tree</summary>

```{r tree, echo=FALSE, message=FALSE}
if (requireNamespace("fs", quietly = TRUE)) {
  fs::dir_tree(recurse = 2, regexp = "^[^._]")
} else {
  cat("Install fs package to see directory tree: install.packages('fs')\n")
}
```

</details>

## Contributing

Contributions are welcome! Please:

1. **Report issues** at [GitHub Issues](https://github.com/JohnGavin/micromort/issues)
2. **Submit PRs** following the [tidyverse style guide](https://style.tidyverse.org/)
3. **Add data** - New risk sources welcome! Include:
   - Source URL with citation
   - Units (micromorts per event OR microlives per day)
   - Period specification (per jump, per day, per year, etc.)

### Development Setup

Requires [Nix](https://nixos.org/) (the `./default.sh` script builds a reproducible R environment via Nix):

```bash
# Clone and enter Nix environment
git clone https://github.com/JohnGavin/micromort.git
cd micromort
./default.sh   # Requires Nix — builds reproducible R env with all dependencies

# Run tests
Rscript -e "devtools::test()"

# Check package
Rscript -e "devtools::check()"
```

## Glossary {#glossary}

| Acronym | Full Name | Definition |
|---------|-----------|------------|
| DALY | Disability-Adjusted Life Year | Disease burden = [YLL](#glossary) + [YLD](#glossary). 1 DALY = 1 year of healthy life lost. |
| LLE | Loss of Life Expectancy | Expected lifespan reduction from a risk, in minutes. See `lle()`. |
| QALD | Quality-Adjusted Life Day | 1 day of perfect health. Useful for short-duration conditions. |
| QALY | Quality-Adjusted Life Year | 1 year of perfect health. Used in healthcare cost-effectiveness. |
| VSL | Value of Statistical Life | Monetary value society places on preventing one death (~$10M USD). See `value_of_micromort()`. |
| YLD | Years Lived with Disability | Morbidity component of [DALY](#glossary). Time spent in impaired health. |
| YLL | Years of Life Lost | Mortality component of [DALY](#glossary). Premature death relative to standard life expectancy. |

## References

- Howard RA (1980). "On Making Life and Death Decisions." *Societal Risk Assessment*.
- Spiegelhalter D (2012). "Using speed of ageing and 'microlives'." BMJ 345:e8223. [DOI: 10.1136/bmj.e8223](https://doi.org/10.1136/bmj.e8223)
- Blastland M, Spiegelhalter D (2013). *The Norm Chronicles: Stories and Numbers About Danger*.

## License

MIT © [John Gavin](https://github.com/JohnGavin)

See [LICENSE](LICENSE) for details.
