---
title: "micromort: Curated Micromort and Microlife Risk Datasets"
format: gfm
---

<!-- README.md is generated from README.qmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
devtools::load_all()
library(ggplot2)
```

# micromort

A **data package** providing curated datasets of micromort (acute risk) and microlife (chronic risk) values from authoritative sources including Wikipedia, CDC MMWR, and academic literature.

## Features

- **62 acute risks** measured in micromorts (one-in-a-million death chance)
- **22 chronic factors** measured in microlives (30 minutes of life expectancy)
- **14 authoritative sources** with full provenance tracking
- **Parquet datasets** for cross-language compatibility (R, Python, Arrow)
- **Plumber REST API** for programmatic access
- **Interactive dashboard** for data exploration
- **targets pipeline** for reproducible data updates

## Installation

```r
# Install from GitHub
# install.packages("devtools")
devtools::install_github("JohnGavin/micromort")
```

### Nix Users

This project provides a reproducible Nix environment:

```bash
# Enter project shell
./default.sh

# Start R
R
```

## Quick Start

### Load the Datasets

```{r load-data}
library(micromort)

# Load acute risks (micromorts)
acute <- load_acute_risks()
nrow(acute)

# Load chronic risks (microlives)
chronic <- load_chronic_risks()
nrow(chronic)

# Load source registry
sources <- load_sources()
nrow(sources)
```

### Explore Acute Risks

```{r acute}
# Top 10 riskiest activities
acute |>
  dplyr::select(activity, micromorts, category) |>
  head(10)
```

### Explore Chronic Factors

```{r chronic}
# Factors that reduce life expectancy
chronic |>
  dplyr::filter(direction == "loss") |>
  dplyr::select(factor, microlives_per_day, category) |>
  head(10)
```

### Visualize Risks

```{r plot}
plot_risks()
```

## Analysis Functions

### Compare Lifestyle Interventions

```{r compare}
# Compare quitting smoking vs losing weight
compare_interventions(list(
  "Quit 10 cigarettes/day" = list(factor = "Smoking 10 cigarettes", change = -1),
  "Lose 5kg" = list(factor = "Being 5 kg overweight", change = -1)
))
```

### Calculate Baseline Risk by Age

```{r hazard}
# Daily baseline risk at age 35
daily_hazard_rate(35)
```

### Lifestyle Tradeoffs

```{r tradeoff}
# How much exercise offsets smoking?
lifestyle_tradeoff("Smoking 2 cigarettes", "20 min moderate exercise")
```

## API Access

Launch the REST API for programmatic access:

```r
launch_api()
# API available at http://localhost:8080
# Swagger docs at http://localhost:8080/__docs__/
```

Endpoints:
- `GET /v1/acute` - Acute risks dataset
- `GET /v1/chronic` - Chronic risks dataset
- `GET /v1/sources` - Source registry
- `GET /v1/hazard?age=35` - Daily hazard rate

## Interactive Dashboard

Launch the Shiny dashboard:

```r
launch_dashboard()
```

## Project Structure

```{r tree, echo=FALSE, results='asis'}
cat("```\n")
cat("micromort/\n")
cat("├── R/                    # Package functions\n")
cat("│   ├── data.R            # Dataset loaders\n")
cat("│   ├── models.R          # Analysis functions\n")
cat("│   ├── api.R             # API launcher\n")
cat("│   └── tar_plans/        # Pipeline plans\n")
cat("├── box/                  # box modules\n")
cat("│   ├── data/             # Data loading\n")
cat("│   ├── models/           # Analysis\n")
cat("│   └── api/              # API endpoints\n")
cat("├── data-raw/             # Source data\n")
cat("│   └── sources/          # CSV source files\n")
cat("├── inst/extdata/         # Parquet datasets\n")
cat("│   ├── acute_risks.parquet\n")
cat("│   ├── chronic_risks.parquet\n")
cat("│   └── risk_sources.parquet\n")
cat("└── _targets.R            # Pipeline orchestrator\n")
cat("```\n")
```

## Data Sources

| Source | Type | Data |
|--------|------|------|
| Wikipedia: Micromort | Encyclopedia | ~50 acute risks |
| Wikipedia: Microlife | Encyclopedia | ~20 chronic risks |
| micromorts.rip | Database | ~45 acute risks |
| CDC MMWR | Government | COVID vaccine data |
| Spiegelhalter (2012) BMJ | Academic | Microlife framework |

## Concepts

- **Micromort:** A one-in-a-million chance of death. Used for acute risks.
- **Microlife:** 30 minutes of life expectancy gained or lost. Used for chronic risks.
- **Conversion:** 1 micromort ≈ 0.7 microlives (assuming 40 years remaining)

## References

- Howard RA (1980). "On Making Life and Death Decisions."
- Spiegelhalter D (2012). BMJ 2012;345:e8223. [DOI: 10.1136/bmj.e8223](https://doi.org/10.1136/bmj.e8223)

## License

MIT
