---
title: "The Invisible Risk"
subtitle: "Why chronic daily disease dwarfs the dangers you worry about"
format:
  closeread-html:
    theme:
      dark: darkly
      light: flatly
    code-fold: true
    code-summary: "Show code"
---

```{=html}
<style>
/* Default: dark theme for closeread scrollytelling */
body { background-color: #1a1a1a; color: #e0e0e0; }
.cr-section { background-color: #1a1a1a; }
.sticky-col { background-color: #1a1a1a; }
table { color: #e0e0e0; }
th, td { border-color: #444 !important; }
h2, h3 { color: #ffffff; }
strong { color: #ffffff; }
a { color: #6ea8fe; }
/* Light mode override for users without dark extensions */
@media (prefers-color-scheme: light) {
  body { background-color: #ffffff; color: #212529; }
  .cr-section { background-color: #ffffff; }
  .sticky-col { background-color: #ffffff; }
  table { color: #212529; }
  th, td { border-color: #dee2e6 !important; }
  h2, h3 { color: #212529; }
  strong { color: #212529; }
  a { color: #0d6efd; }
}
</style>
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "")

if (requireNamespace("micromort", quietly = TRUE)) {
  library(micromort)
} else {
  pkgload::load_all(quiet = TRUE)
}

# Shared safe_tar_read with RDS fallback (see inst/vignette_utils.R)
source(file.path(tryCatch(rprojroot::find_root(rprojroot::is_r_package),
  error = function(e) "."), "inst", "vignette_utils.R"))
```

:::{.cr-section}

## The invisible risk

A heroin dose carries **30 micromorts**. Base jumping: **430**. These feel dangerous because they're *chosen* and *visible*.

But what if the bigger risk is something you can't see, can't feel, and can't opt out of? @cr-equiv-table

:::{#cr-equiv-table .sticky}

```{r equiv-table}
#| echo: false
show_target("vig_chronic_acute_heroin_table",
  caption = "Days of chronic disease risk equivalent to one acute event (micromorts). 9 days of UK CVD risk equals one heroin dose (30 mm). Source: common_risks() and chronic_disease_risks().")
```

:::

[**9 days.** That's how long it takes for background cardiovascular disease risk in the UK to accumulate the same 30 micromorts as a single dose of heroin.]{highlight="cr-equiv-table"}

[Every 9 days, your heart silently rolls the same dice as a heroin user — but nobody stages an intervention.]{highlight="cr-equiv-table"}

:::


:::{.cr-section}

## The annual picture

Over a full year, chronic disease risk dwarfs any one-off activity. Here's the UK annual chronic disease burden alongside the acute activities that make headlines: @cr-waterfall

:::{#cr-waterfall .sticky}

### UK annual chronic disease risk (micromorts/year)

```{r waterfall-chronic}
#| echo: false
show_target("vig_chronic_acute_waterfall_chronic",
  caption = "UK annual chronic disease micromorts. Cancer (~1,400 mm/yr) and CVD (~1,200 mm/yr) together equal climbing the Matterhorn annually. Source: chronic_disease_risks(country='GBR').")
```

### Acute one-off risks for comparison

```{r waterfall-acute}
#| echo: false
show_target("vig_chronic_acute_waterfall_acute",
  caption = "Acute one-off risks for comparison (micromorts per event). Skydiving (8 mm) and marathons (7 mm) are dwarfed by annual chronic disease totals. Source: common_risks().")
```

:::

[UK cancer alone accumulates over **1,400 micromorts per year**. [Cardiovascular disease (CVD)](https://www.who.int/health-topics/cardiovascular-diseases) adds another **1,200**. Together, that's the equivalent of climbing the Matterhorn — every single year.]{highlight="cr-waterfall"}

[Yet we fear the mountain, not the mundane.]{highlight="cr-waterfall"}

:::


:::{.cr-section}

## Country matters

The chronic burden is not fixed — it depends enormously on where you live. Switching from a UK to Nigerian profile changes everything: @cr-country

:::{#cr-country .sticky}

```{r country-shift}
#| echo: false
show_target("vig_chronic_acute_country_display",
  caption = "Daily disease micromorts: UK vs Nigeria. Diarrheal disease 44x higher in Nigeria; cardiovascular disease (CVD) doubles. Geography is the dominant confounder in chronic risk. Source: common_risks() with country profiles.")
```

:::

[**Diarrheal disease** jumps from 0.02 mm/day in the UK to 0.88 mm/day in Nigeria — a 44x difference driven by clean water and sanitation infrastructure.]{highlight="cr-country"}

[**<a href="https://www.who.int/health-topics/cardiovascular-diseases"><abbr title="Cardiovascular disease: diseases of the heart and blood vessels">CVD</abbr></a>** doubles from 3.2 to 7.3 mm/day. The same organ, the same biology, but vastly different outcomes depending on healthcare access, diet, and environmental factors.]{highlight="cr-country"}

[These are the confounders that population-average micromort tables hide. See the [confounding vignette](confounding.html) for the full geographic and demographic analysis.]{highlight="cr-country"}

:::


:::{.cr-section}

## The takeaway

The risks we *choose* (skydiving, motorcycling, mountaineering) are visible, voluntary, and bounded. The risks we *inherit* (cardiovascular disease, cancer, respiratory infections) are invisible, involuntary, and cumulative.

The micromort framework makes them commensurable — you can compare a skydive to a week of heart disease, a flight to a month of air pollution, a surgery to a year of background cancer risk.

Explore the data yourself:

- `common_risks(profile = list(country = "UK"))` — your country's disease burden
- `geography_quiz_pairs()` — quiz comparing risks across countries
- `risk_equivalence()` — formal equivalence calculations

:::

<details>
<summary>Session info (click to expand)</summary>

```{r session-info, eval=TRUE, echo=FALSE}
sessionInfo()
```

</details>

```{r build-info}
#| echo: false
#| results: asis
cat(safe_tar_read("vig_build_info") %||% "")
```
