---
title: "The Invisible Risk"
subtitle: "Why chronic daily disease dwarfs the dangers you worry about"
format:
  closeread-html:
    theme:
      dark: darkly
      light: flatly
    code-fold: true
    code-summary: "Show code"
---

```{=html}
<style>
body { background-color: #1a1a1a; color: #e0e0e0; }
.cr-section { background-color: #1a1a1a; }
.sticky-col { background-color: #1a1a1a; }
table { color: #e0e0e0; }
th, td { border-color: #444 !important; }
h2, h3 { color: #ffffff; }
strong { color: #ffffff; }
a { color: #6ea8fe; }
</style>
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")

if (requireNamespace("micromort", quietly = TRUE)) {
  library(micromort)
} else {
  pkgload::load_all(quiet = TRUE)
}

safe_tar_read <- function(name) {
  store <- Sys.getenv("TARGETS_STORE", unset = NA)
  if (is.na(store)) {
    store <- tryCatch(
      file.path(rprojroot::find_root(rprojroot::is_r_package), "_targets"),
      error = function(e) "_targets"
    )
  }
  tryCatch(
    targets::tar_read_raw(name, store = store),
    error = function(e) {
      message("Target '", name, "' not found. Run tar_make() first.")
      NULL
    }
  )
}
```

:::{.cr-section}

## One skydive

A heroin dose carries **30 micromorts**. Base jumping: **430**. These feel dangerous because they're *chosen* and *visible*.

But what if the bigger risk is something you can't see, can't feel, and can't opt out of? @cr-equiv-table

:::{#cr-equiv-table .sticky}

```{r equiv-table}
#| echo: false
equiv <- safe_tar_read("vig_chronic_acute_equivalences")
if (!is.null(equiv)) {
  # Show heroin dose vs UK chronic risks (compelling comparison)
  heroin <- equiv[equiv$activity == "Heroin use (per dose)", ]
  if (nrow(heroin) > 0) {
    heroin |>
      dplyr::select(`Chronic daily risk` = daily_activity,
                     `Daily mm` = daily_mm,
                     `Days to equal one heroin dose` = days_equivalent) |>
      knitr::kable(digits = 1)
  }
} else {
  cat("*Run tar_make() to see equivalence data.*")
}
```

:::

[**9 days.** That's how long it takes for background cardiovascular disease risk in the UK to accumulate the same 30 micromorts as a single dose of heroin.]{highlight="cr-equiv-table"}

[Every 9 days, your heart silently rolls the same dice as a heroin user — but nobody stages an intervention.]{highlight="cr-equiv-table"}

:::


:::{.cr-section}

## The annual picture

Over a full year, chronic disease risk dwarfs any one-off activity. Here's the UK annual chronic disease burden alongside the acute activities that make headlines: @cr-waterfall

:::{#cr-waterfall .sticky}

```{r waterfall}
#| echo: false
wf <- safe_tar_read("vig_chronic_acute_waterfall")
if (!is.null(wf)) {
  cat("### UK annual chronic disease risk (micromorts/year)\n\n")
  wf$chronic_annual |> knitr::kable(digits = 1)
  cat("\n### Acute one-off risks for comparison\n\n")
  wf$acute_comparison |> knitr::kable(digits = 1)
} else {
  cat("*Run tar_make() to see waterfall data.*")
}
```

:::

[UK cancer alone accumulates over **1,400 micromorts per year**. CVD adds another **1,200**. Together, that's the equivalent of climbing the Matterhorn — every single year.]{highlight="cr-waterfall"}

[Yet we fear the mountain, not the mundane.]{highlight="cr-waterfall"}

:::


:::{.cr-section}

## Country matters

The chronic burden is not fixed — it depends enormously on where you live. Switching from a UK to Nigerian profile changes everything: @cr-country

:::{#cr-country .sticky}

```{r country-shift}
#| echo: false
cs <- safe_tar_read("vig_chronic_acute_country_shift")
if (!is.null(cs)) {
  cs |>
    dplyr::select(Cause = cause,
                   `UK (mm/day)` = uk_mm,
                   `Nigeria (mm/day)` = ng_mm,
                   `Ratio (NG/UK)` = ratio) |>
    knitr::kable(digits = 2)
} else {
  cat("*Run tar_make() to see country comparison.*")
}
```

:::

[**Diarrheal disease** jumps from 0.02 mm/day in the UK to 0.88 mm/day in Nigeria — a 44x difference driven by clean water and sanitation infrastructure.]{highlight="cr-country"}

[**CVD** doubles from 3.2 to 7.3 mm/day. The same organ, the same biology, but vastly different outcomes depending on healthcare access, diet, and environmental factors.]{highlight="cr-country"}

[These are the confounders that population-average micromort tables hide. See the [confounding vignette](confounding.html) for the full geographic and demographic analysis.]{highlight="cr-country"}

:::


:::{.cr-section}

## The takeaway

The risks we *choose* (skydiving, motorcycling, mountaineering) are visible, voluntary, and bounded. The risks we *inherit* (cardiovascular disease, cancer, respiratory infections) are invisible, involuntary, and cumulative.

The micromort framework makes them commensurable — you can compare a skydive to a week of heart disease, a flight to a month of air pollution, a surgery to a year of background cancer risk.

Explore the data yourself:

- `common_risks(profile = list(country = "UK"))` — your country's disease burden
- `geography_quiz_pairs()` — quiz comparing risks across countries
- `risk_equivalence()` — formal equivalence calculations

:::

<details>
<summary>Session info (click to expand)</summary>

```{r session-info, eval=TRUE, echo=FALSE}
sessionInfo()
```

</details>
