---
title: "How Reliable Are These Numbers?"
format: html
vignette: >
  %\VignetteIndexEntry{How Reliable Are These Numbers?}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = FALSE
)

if (requireNamespace("micromort", quietly = TRUE)) {
  library(micromort)
} else {
  pkgload::load_all(quiet = TRUE)
}
library(dplyr)
```

Risk numbers disagree. The WHO and the Institute for Health Metrics and Evaluation (IHME) report malaria deaths as 550,000 and 760,000 respectively — a 38% gap from the *same underlying deaths*. Our World in Data's [Deadliest Animals](https://ourworldindata.org/deadliest-animals) chart is visually compelling, but converting annual death counts to per-encounter micromorts is non-trivial. This vignette documents how we handle that uncertainty.

## 1. Why Risk Numbers Disagree

Three factors drive disagreement between sources:

1. **Numerator uncertainty**: Death attribution varies by coding system (ICD-10 codes, verbal autopsy, hospital records)
2. **Denominator uncertainty**: How many people were *exposed*? A "deaths per year" figure means nothing without knowing the exposure population
3. **Temporal and geographic aggregation**: A global annual average hides enormous regional and seasonal variation

Our inclusion criteria: **traceable numerator** + **defined denominator** + **reproducible calculation**. We reject risks where we cannot identify both the death count *and* the population at risk.

## 2. The Confidence System

Every entry in `atomic_risks()` carries a `confidence` tier:

```{r confidence-table}
#| tbl-cap: "Confidence tiers with examples from the micromort dataset"
ar <- atomic_risks()

tibble::tribble(
  ~Tier, ~Criteria, ~Example, ~`Source type`,
  "**high**", "Peer-reviewed, large-N studies with defined denominators",
  "Medical radiation (NRC dosimetry)", "Regulatory agency",
  "**medium**", "Reputable sources, reasonable denominators, some extrapolation",
  "Wikipedia micromort list, CDC injury data", "Secondary compilation",
  "**low**", "Limited sources, regional uncertainty, or extrapolated denominators",
  "Snake bite in rural Africa (WHO estimate)", "Expert estimate",
  "**estimated**", "Derived by calculation from a model (e.g., LNT for radiation)",
  "Annual cosmic radiation from LNT model", "Model-derived"
) |>
  knitr::kable()
```

### Validation status (new)

Within each confidence tier, we now track how thoroughly the estimate has been cross-checked:

```{r validation-table}
#| tbl-cap: "Validation status levels"
tibble::tribble(
  ~Status, ~Definition, ~`Source count`, ~Example,
  "`single_source`", "One citation, no cross-check", "1",
  "Most legacy entries from Wikipedia/micromorts.rip",
  "`corroborated`", "2+ sources agree within 2x", "2+",
  "Flight risks (Boeing + NCRP + medical literature)",
  "`cross_validated`", "3+ sources, range documented, outliers explained", "3+",
  "(Future: entries with systematic literature review)"
) |>
  knitr::kable()
```

```{r validation-summary}
#| tbl-cap: "Current validation status across all entries"
#| fig-cap: "Distribution of validation status in the dataset"
ar |>
  count(validation_status, confidence) |>
  tidyr::pivot_wider(names_from = validation_status, values_from = n, values_fill = 0) |>
  knitr::kable()
```

## 3. Geographic Conditioning: The Biggest Source of Variation

The same animal encounter can produce dramatically different micromort values depending on location and healthcare access:

```{r geography-table}
#| tbl-cap: "Geographic conditioning: same encounter, different risk"
ar |>
  filter(condition_variable == "geography") |>
  select(activity, micromorts, condition_value, confidence, notes) |>
  arrange(activity, condition_value) |>
  knitr::kable(digits = 1)
```

Snake bite: **0.5 mm** (US, with antivenom) vs **18.5 mm** (rural Africa) — a **37x difference**.
Dog bite: **6.7 mm** (US, with rabies PEP) vs **160 mm** (rabies-endemic, no treatment) — a **24x difference**.

### The hedgeability asymmetry

Geography is a **hedgeable conditional risk** — but only for some people:

- A **tourist** can hedge: choose destination, get travel vaccines, carry antivenom kit, buy travel insurance
- A **resident** cannot hedge: they live there, and may lack healthcare infrastructure, vaccines, or economic choice

This parallels the existing health profile conditioning. A bee sting is 0.03 mm for someone who is not allergic, but **31 mm** for someone with a known allergy — a 1,000x difference. The allergic person can hedge (carry an epinephrine auto-injector), but they cannot eliminate the underlying vulnerability.

```{r health-conditioning}
#| tbl-cap: "Health profile conditioning: same sting, different risk"
ar |>
  filter(condition_variable == "health_profile", grepl("bee|wasp", activity, ignore.case = TRUE)) |>
  select(activity, micromorts, condition_value, hedge_description, hedge_reduction_pct) |>
  knitr::kable(digits = 2)
```

### Using geographic filtering

```r
# Default: returns high-income estimates
common_risks() |> filter(category == "Wildlife")

# Explicitly request low-income geography
common_risks(profile = list(geography = "low_income")) |> filter(category == "Wildlife")

# Combine with health profile
common_risks(profile = list(geography = "low_income", health_profile = "allergic"))
```

## 4. Cross-Validation Methods

We use five methods to assess data reliability:

### Source triangulation

Compare the same risk across independent sources. For wildlife risks, we cross-reference:

- **OWID** annual death counts (numerator)
- **CDC** injury surveillance (US denominator)
- **WHO** fact sheets (global denominator)
- **ISAF** shark attack database (species-specific data)

### Denominator audit

The most common failure mode. Does the source report **both** a numerator (deaths) **and** a denominator (exposures)?

| Animal | Numerator available? | Denominator available? | Included? |
|--------|---------------------|----------------------|-----------|
| Shark | Yes (ISAF) | Yes (~100M swims/yr) | Yes |
| Dog | Yes (CDC, WHO) | Yes (4.5M bites US) | Yes |
| Mosquito | Yes (WHO: 600k+) | No per-encounter rate | **No** |
| Crocodile | Yes (CrocBITE) | No exposure estimate | **No** |

### Temporal stability

Has the number changed significantly across editions of the source? Stable estimates across 5+ years increase confidence.

### Geographic consistency

Do US, UK, and global estimates agree within an order of magnitude? Large discrepancies suggest unmeasured confounders (see [Confounding Variables](confounding.html)).

### Order-of-magnitude test

Is the number physically plausible? A micromort value that implies more deaths than the population can support is a red flag.

## 5. Worked Example: Animal Risks from OWID

Our World in Data reports annual deaths by animal. Converting to per-encounter micromorts requires:

$$\text{micromorts} = \frac{\text{deaths per year}}{\text{encounters per year}} \times 10^6$$

```{r owid-conversion}
#| tbl-cap: "Converting OWID annual counts to per-encounter micromorts"
tibble::tribble(
  ~Animal, ~`Annual deaths (approx)`, ~`Encounters/yr (approx)`,
  ~`Micromorts`, ~`Source for denominator`, ~`In dataset?`,
  "Shark", "~6 (US)", "~100M ocean swims", "0.06", "ISAF", "Yes",
  "Dog (US)", "~30", "~4.5M bites", "6.7", "CDC", "Yes",
  "Bee/wasp (US)", "~62", "~2M stings", "0.03", "CDC", "Yes",
  "Snake (US)", "~5", "~10,000 bites", "0.5", "CDC", "Yes",
  "Snake (Africa)", "~100,000", "~5.4M bites", "18.5", "WHO/Lancet", "Yes",
  "Mosquito", "~600,000+", "Unknown per-bite", "—", "—", "**No**",
  "Crocodile", "~1,000", "Unknown", "—", "—", "**No**",
  "Elephant", "~500", "Unknown", "—", "—", "**No**"
) |>
  knitr::kable()
```

Mosquito, crocodile, and elephant fail our inclusion criteria: there is no defensible per-encounter denominator. Mosquito bites are ubiquitous in endemic regions, making a per-bite risk meaningless. We cite OWID for context but do not include these as micromort entries.

## 6. Estimate Ranges

For wildlife entries, we document plausible ranges reflecting source disagreement:

```{r ranges}
#| tbl-cap: "Estimate ranges for wildlife entries"
ar |>
  filter(category == "Wildlife", !is.na(estimate_range)) |>
  select(activity, micromorts, estimate_range, source_count, validation_status) |>
  knitr::kable(digits = 2)
```

The range reflects uncertainty in both the numerator (death counts vary by year and reporting) and denominator (exposure estimates are often rough). The point estimate is our best central value; the range brackets the plausible minimum and maximum.

## 7. What You Can Contribute

If you find a better source for an existing entry, or want to propose a new risk:
open an issue at [github.com/johngavin/micromort](https://github.com/johngavin/micromort/issues) with:

1. **Numerator**: Death count and source citation
2. **Denominator**: Exposure count and source citation
3. **Geography/condition**: Does the estimate apply globally, or to a specific population?
4. **Time period**: When was the data collected?

Entries start at `validation_status = "single_source"` and get upgraded as more sources confirm them.
