---
title: "Details"
description: "Retired Shinylive quiz links, source locations, and site build info for the micromort package — consolidated off the quiz pages so they don't interrupt quiz-taking."
format:
  html:
    code-fold: true
    code-summary: "Show code"
vignette: >
  %\VignetteIndexEntry{Details}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = ""
)

if (requireNamespace("micromort", quietly = TRUE)) {
  library(micromort)
} else {
  pkgload::load_all(quiet = TRUE)
}
```

This page collects information that used to sit at the bottom of the quiz
pages: the Shinylive app links, how to view a page's source, and site build
info. It was moved here because on the quiz pages that content sat directly
below the interactive quiz area, so it was encountered on every scroll while
answering questions — see
[issue #142](https://github.com/JohnGavin/micromort/issues/142) for the full
history.

::: {.panel-tabset}

## Retired: Shinylive quizzes

The three quizzes each used to have a second, full Shiny app version
("Shinylive") alongside the instant (static JS) version below, offering
score submission and percentile ranking. Those Shinylive versions were
retired on 2026-09-05: an audit found their only remaining capability not
already in the instant version was a small played-N-days-in-a-row streak
indicator, which has since been ported to the instant quizzes. Retiring the
Shinylive copies removed roughly 180MB of duplicate WASM bundles from this
site and stopped every quiz-page fix from needing two parallel
implementations kept in sync.

The retired pages are still available in the repository history at the
[`archive/shinylive-quizzes-2026-09-05`](https://github.com/JohnGavin/micromort/tree/archive/shinylive-quizzes-2026-09-05)
tag.

| Quiz | Instant version |
|---|---|
| Which lifestyle event is more likely to kill you? (micromorts) | [micromort-quiz.html](micromort-quiz.html) |
| Which daily habit has a bigger effect on your lifespan? (microlives) | [microlife-quiz.html](microlife-quiz.html) |
| Risk ranking | [risk-ranking-quiz.html](risk-ranking-quiz.html) |

## Source

Every article page on this site has its own auto-generated "Source:" line
naming the `.qmd` file it was rendered from — that line is inherently
per-page (pkgdown attaches it individually to each article), so it cannot be
consolidated into a single link here.

To view the source of the page you're currently reading, look for the
"Source:" line near the top of that page, or browse the source directly:

- [vignettes/](https://github.com/JohnGavin/micromort/tree/main/vignettes) —
  every quiz, article, and dashboard `.qmd` source file
- [R/](https://github.com/JohnGavin/micromort/tree/main/R) — package
  functions and data pipeline code
- [Full repository](https://github.com/JohnGavin/micromort)

## Site info

```{r site-info, echo=FALSE, results='asis'}
pkg_version <- tryCatch(
  as.character(utils::packageVersion("micromort")),
  error = function(e) "unknown"
)
cat(sprintf(
  "- **Package:** micromort %s\n- **Site built with:** [pkgdown](https://pkgdown.r-lib.org/)\n- **Commit history:** [github.com/JohnGavin/micromort/commits/main](https://github.com/JohnGavin/micromort/commits/main)\n",
  pkg_version
))
```

:::
