---
title: "Introduction to Micromorts and Risk Visualization"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to Micromorts and Risk Visualization}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(micromort)
library(ggplot2)
library(dplyr)
library(DT)
```

This vignette introduces the **micromort** package, which provides tools for understanding and visualizing risks.

## 1. Micromorts (Acute Risk)

A **micromort** is a unit of risk representing a one-in-a-million chance of death. More precisely, it's a **microprobability** of death — a one-in-a-million chance of a specific event (death) occurring.

### Definition

| Term | Definition | Example |
|------|------------|---------|
| **Microprobability** | 1-in-a-million chance of any event | 1 micromort = microprobability of death |
| **Micromort** | 1-in-a-million chance of death, per event | Skydiving: 8 micromorts per jump |

### Comparing Risks: Period Matters!

**CRITICAL:** When comparing micromort values, ensure the **period is the same**. For example:

| Activity | Micromorts | Period | Comparable? |
|----------|------------|--------|-------------|
| Scuba diving | 5 | per dive | ✓ Per-event |
| Scuba diving | 164 | per year | ✗ Per-year (assumes ~33 dives/year) |
| Skydiving | 8 | per jump | ✓ Per-event |

The "per year" figure (164) conflates frequency with risk-per-event. A diver doing 5 dives/year vs 50 dives/year faces very different annual risk.

### Conditional Risks

Many activities have **selection effects**. For example:

*   **Marathon running (7 micromorts):** Runners are self-selected for fitness. This low figure reflects the health of participants, not the risk to an average person attempting a marathon.
*   **Motorcycle riding (10 micromorts/60 miles):** Experienced riders face lower risk than novices, but the quoted figure is an average.

These figures answer: "Given that someone completed this activity, what was their death risk?" Not: "What would happen if a random person attempted this?"

```{r}
# 1 in 10,000 chance of death = 100 micromorts
prob <- 1/10000
as_micromort(prob)

# Compare common risks (sortable table - click column headers)
risks <- common_risks()
DT::datatable(
  risks,
  caption = "Common risks in micromorts (click column headers to sort, use search box to filter)",
  filter = "top",
  options = list(pageLength = 10, scrollX = TRUE),
  rownames = FALSE
)
```

### Visualizing Risks

Using [`plot_risks()`](../reference/plot_risks.html), we can see the relative magnitude of different activities on a logarithmic scale. The plot is split into COVID-19 and Other risks to make comparisons easier:
```{r, fig.width=8, fig.height=14}
# Default: faceted by COVID-19 vs Other
plot_risks()
```

#### Interactive Version

For interactive exploration with hover details and category filtering, use [`plot_risks_interactive()`](../reference/plot_risks_interactive.html):

```{r}
# plotly version with dropdown filter and hover details
plot_risks_interactive()
```

## 2. Microlives (Chronic Risk)

While micromorts measure sudden death, **microlives** measure the impact of chronic habits on your life expectancy. A microlife represents a 30-minute change in life expectancy.

### Living at Different Speeds

A useful way to think about microlives: we all "use up" **48 microlives per day** just by living (24 hours = 48 × 30 minutes). Unhealthy habits accelerate this consumption, while healthy habits slow it down.

A smoker who smokes 20 cigarettes per day uses up an additional 10 microlives, which can be interpreted as **rushing towards death at 29 hours per day** instead of 24. Conversely, someone with excellent lifestyle habits might effectively live at only 22 hours per day.

**Healthcare bonus:** Modern healthcare and healthier lifestyles give us a "payback" of approximately **12 microlives per day** — our expected death is moving away from us even as we age.

### Common Chronic Risks

| Factor | Microlives/day | Interpretation |
|--------|----------------|----------------|
| Smoking 1 cigarette | -1 | Lose 30 min life expectancy |
| Being 5kg overweight | -1 | Lose 30 min/day |
| 20 min moderate exercise | +2 | Gain 60 min life expectancy |
| 2+ hours TV daily | -1 | Sedentary behavior |
| 5+ servings fruit/veg | +2 | Healthy diet |

```{r}
# as_microlife() converts minutes of life expectancy change to microlives
# Unit: 1 microlife = 30 minutes of life expectancy change PER DAY
# Sign: negative = loss, positive = gain

# Heavy smoker (20 cigarettes/day × 30 mins each = 600 mins LOST)
# They're living at 24 + (20×0.5) = 29 hours/day towards death!
as_microlife(-20 * 30)  # = -20 microlives/day (life lost)

# Moderate exercise (20 mins exercise → 60 mins life GAINED)
as_microlife(60)        # = +2 microlives/day (life gained)

# Being 5kg overweight costs 30 mins per day
as_microlife(-30)       # = -1 microlife/day (life lost)

# Exercise can offset smoking:
# 2 cigarettes = -60 mins, 20 min exercise = +60 mins → net zero
as_microlife(-60) + as_microlife(60)  # = 0 (they cancel out)
```

## 3. Relationship Between Micromorts and Microlives

### Theoretical Conversion

Micromorts (acute, per-event risk) and microlives (chronic, per-day attrition) measure different phenomena, but can be approximately converted using expected value theory.

**Key relationship:** 1 micromort ≈ 0.7 microlives

**Assumptions for this conversion:**
1. Remaining life expectancy = 40 years (adjust for actual age)
2. Death occurs immediately upon the event (worst case)
3. Linear approximation (valid for small probabilities)

**Mathematical derivation:**

```{r}
# UNITS: 1 micromort = 1-in-a-million probability of IMMEDIATE DEATH per EVENT
# Question: What is the expected life lost from ONE micromort event?

remaining_years <- 40  # Assumed remaining life expectancy (years)
prob_death <- 1e-6     # 1 micromort = 1/1,000,000 death probability

# Expected life lost = probability × life forfeited if death occurs
# If the event kills you, you lose all remaining years
life_lost_years <- prob_death * remaining_years
cat("Expected life lost (years):", life_lost_years, "\n")
cat("  = 1e-6 × 40 = 4e-5 years\n\n")

# Convert years to minutes: 1 year = 365.25 × 24 × 60 = 525,960 minutes
life_lost_minutes <- life_lost_years * 365.25 * 24 * 60
cat("Expected life lost (minutes):", round(life_lost_minutes, 1), "\n")
cat("  = 4e-5 × 525,960 ≈ 21 minutes\n\n")

# 1 microlife = 30 minutes of life expectancy change
# So 21 minutes ≈ 0.7 microlives
microlives_equivalent <- life_lost_minutes / 30
cat("Microlives equivalent:", round(microlives_equivalent, 2), "\n")
```

**Why `common_risks()` uses microlives = micromorts × 0.7:**

The conversion allows comparing a single risky event (like one skydive at 8 micromorts) to chronic daily habits (like smoking at -10 microlives/day). However, **the approximation breaks down when:**

1. **Remaining life expectancy differs** from 40 years (a 20-year-old loses more per micromort than an 80-year-old)
2. **The risk is not immediate death** (injuries, disabilities are not captured)
3. **Repeated exposures** compound non-linearly

### Unit Definitions (Summary)

| Metric | Unit Definition | Scope | Sign |
|--------|-----------------|-------|------|
| **Micromort** | 1-in-a-million probability of death | Per discrete event (1 surgery, 1 flight) | Always ≥ 0 |
| **Microlife** | 30 minutes of life expectancy change | Per day of exposure/habit | + = gain, − = loss |

### When to Use Each

*   **Use micromorts** for discrete, short-duration events with binary outcomes (death or survival): surgery, skydiving, a single car trip
*   **Use microlives** for chronic, daily habits that accumulate over a lifetime: smoking, exercise, diet
*   **Convert between them** for policy decisions, but state your assumptions (age, remaining life expectancy)

## 4. Value of Statistical Life (VSL)

The **Value of a Statistical Life (VSL)** is the monetary value used to justify safety spending. It is NOT the value of an individual life, but the aggregate willingness to pay for small risk reductions.

### US Valuation

Example: If a safety feature costs $50 and saves 1 life in 100,000 people (10 micromorts), is it worth it?
Cost per micromort saved = $50 / 10 = $5.
If VSL = $10M, then 1 micromort = $10. Since $5 < $10, it is cost-effective.

```{r}
# Standard US VSL of $10M implies $10 per micromort
value_of_micromort(vsl = 10000000)
```

### UK Valuation: Micromorts ≈ Microlives

Interestingly, two UK government agencies arrive at similar valuations for micromorts and microlives:

| Agency | Metric | Value | Per Unit |
|--------|--------|-------|----------|
| **NICE** (NHS) | 1 QALY | ~£30,000 | **£1.70 per microlife** |
| **Dept of Transport** | Value of Statistical Life | £1,600,000 | **£1.60 per micromort** |

This near-equivalence (£1.60 ≈ £1.70) provides empirical support for the theoretical conversion: **1 micromort ≈ 1 microlife** in policy terms.

```{r}
# UK Department of Transport VSL: £1.6M → £1.60 per micromort
value_of_micromort(vsl = 1600000)

# Compare: NICE prices a microlife at ~£1.70
# (£30,000 per QALY ÷ 365 days ÷ 48 microlives/day ≈ £1.71)
nice_per_microlife <- 30000 / 365 / 48
round(nice_per_microlife, 2)
```

This consistency suggests that policy decisions affecting acute risks (transport safety) and chronic risks (healthcare interventions) can be compared on a common scale.

## 5. Loss of Life Expectancy (LLE)

**LLE** estimates the average time lost from a lifespan due to a specific risk. For a 1-in-a-million risk (1 micromort), the LLE is tiny.

```{r}
# Loss of life expectancy from 1 micromort (assuming 40 years remaining)
lle_minutes <- lle(prob = 1/1e6, life_expectancy = 40)
print(lle_minutes)
# Result is in minutes. 1 micromort ~ 21 minutes lost?
# No, 1 micromort = 1e-6 * 40 years = 40e-6 years = ~21 minutes
# Wait, check calculation:
# 40 years * 365.25 days * 24 hours * 60 minutes = ~21 million minutes
# 1e-6 * 21 million = ~21 minutes
```

## 6. Complementary Metrics: QALY, DALY, and Morbidity

Micromorts and microlives focus on mortality. But many conditions (like the common cold) cause significant quality of life loss without being fatal. Complementary metrics capture this morbidity burden.

### QALY (Quality-Adjusted Life Year)

Measures years of life adjusted for quality. **1 QALY = 1 year of perfect health.**

*   Health states are weighted 0 (death) to 1 (perfect health)
*   A year with chronic pain at 0.7 quality = 0.7 QALYs
*   Used to assess cost-effectiveness of medical interventions (e.g., £20,000-30,000 per QALY threshold in UK)

### DALY (Disability-Adjusted Life Years)

Measures disease burden as the sum of:

*   **YLL (Years of Life Lost):** From premature mortality
*   **YLD (Years Lived with Disability):** From morbidity, weighted by disability severity

**Formula:** `DALY = YLL + YLD`

For fatal diseases like COVID-19, YLL dominates. For non-fatal conditions like the common cold, YLD dominates.

### Capturing Non-Fatal Burden: The "Microburden" Concept

The common cold doesn't kill many people, but millions of work/school days are lost annually. How do we capture this in a microlife-like metric?

**Proposed "Microburden":** Population-level quality of life lost per million person-days.

```{r}
# Example: Common cold burden calculation
# Disability weight for common cold: ~0.006 (WHO GBD)
# Average duration: 5-7 days
# Episodes per person per year: ~2-3

cold_disability_weight <- 0.006
cold_duration_days <- 6
episodes_per_year <- 2.5

# QALD (Quality-Adjusted Life Days) lost per episode
qald_per_episode <- cold_disability_weight * cold_duration_days

# Annual burden per person (in days of perfect health equivalent)
annual_burden_days <- qald_per_episode * episodes_per_year
cat("Annual quality burden (QALD/person):", round(annual_burden_days, 4), "\n")

# For 1 million people
population <- 1e6
total_qald_lost <- annual_burden_days * population
cat("Population burden (QALD/million):", format(total_qald_lost, big.mark = ","), "\n")
```

### Comparing Metrics

| Metric | Unit Definition | Scope | Sign | Best For |
|--------|-----------------|-------|------|----------|
| **Micromort** | 1/1,000,000 death probability | Per discrete event (surgery, flight, climb) | ≥ 0 (probability) | Comparing single risky activities |
| **Microlife** | 30 min life expectancy change | Per day of chronic exposure | + gain / − loss | Daily lifestyle interventions |
| **QALY** | 1 year at perfect health (quality=1.0) | Per treatment/intervention | ≥ 0 | Cost-effectiveness in healthcare |
| **DALY** | 1 year lost to disease (YLL + YLD) | Per condition/population | ≥ 0 (burden) | Global health prioritization |
| **QALD** | 1 day at perfect health | Per illness episode | ≥ 0 | Short-term morbidity (colds, flu) |

### References

*   Spiegelhalter D (2012). "Using speed of ageing and 'microlives'." BMJ 2012;345:e8223. [doi:10.1136/bmj.e8223](https://doi.org/10.1136/bmj.e8223)
*   Spiegelhalter D (2012). "Understanding uncertainty: Microlives." Plus Magazine. [plus.maths.org](https://plus.maths.org/content/understanding-uncertainty-microlives)
*   WHO Global Burden of Disease: [ghdx.healthdata.org](https://ghdx.healthdata.org/)
*   NICE Methods Guide: [nice.org.uk](https://www.nice.org.uk/process/pmg9/chapter/the-reference-case)

## 7. Conditional Risks: Cancer, Vaccination, and Risk Hedging

### Cancer Risks by Type and Sex

The [`cancer_risks()`](../reference/cancer_risks.html) function provides mortality data stratified by cancer type, sex, and age group:

```{r}
# Top 3 cancers by mortality for each sex
cancer_risks() |>
  dplyr::filter(age_group == "All ages", sex != "Both") |>
  dplyr::group_by(sex) |>
  dplyr::slice_min(rank_by_sex, n = 3) |>
  dplyr::select(cancer_type, sex, deaths_per_100k, micromorts_per_year, family_history_rr) |>
  DT::datatable(
    caption = "Top 3 cancers by mortality rate (click to sort)",
    options = list(pageLength = 6, dom = "t"),
    rownames = FALSE
  )
```

**Family history impact:** The `family_history_rr` column shows relative risk increase with a first-degree relative's diagnosis. For example, prostate cancer risk increases 2.5× with family history.

```{r}
# Compare risk with vs without family history (male, age 50-64)
cancer_risks() |>
  dplyr::filter(sex == "Male", age_group == "50-64") |>
  dplyr::select(cancer_type, micromorts_per_year, micromorts_with_family_history) |>
  dplyr::mutate(
    increase_mm = micromorts_with_family_history - micromorts_per_year
  ) |>
  DT::datatable(
    caption = "Cancer risk with family history (Male, 50-64 years)",
    options = list(pageLength = 5, dom = "t"),
    rownames = FALSE
  )
```

### Vaccination Risk Reduction

The [`vaccination_risks()`](../reference/vaccination_risks.html) function quantifies micromorts avoided through vaccination:

```{r}
# Childhood vaccination impact by country
vaccination_risks() |>
  dplyr::filter(age_group == "0-5", grepl("Complete", vaccine_schedule)) |>
  dplyr::select(country, mortality_reduction_pct, micromorts_avoided_per_year, annual_life_days_gained) |>
  DT::datatable(
    caption = "Complete childhood vaccination schedule impact",
    options = list(pageLength = 5, dom = "t"),
    rownames = FALSE
  )
```

```{r}
# Adult vaccination benefits (age 65+)
vaccination_risks() |>
  dplyr::filter(age_group == "65+", country == "US") |>
  dplyr::select(vaccine_schedule, micromorts_avoided_per_year, microlives_gained_per_day) |>
  DT::datatable(
    caption = "Adult vaccination benefits (US, age 65+)",
    options = list(pageLength = 5, dom = "t"),
    rownames = FALSE
  )
```

### Hedged vs Unhedged: Optimal Lifestyle Comparison

The [`conditional_risk()`](../reference/conditional_risk.html) function compares risk factors between optimal ("hedged") and suboptimal ("unhedged") states:

```{r}
# Cardiovascular risk hedging
conditional_risk("cardiovascular") |>
  dplyr::select(risk_factor, unhedged_state, hedged_state, microlives_gained, annual_days_gained) |>
  DT::datatable(
    caption = "Cardiovascular hedging: microlives gained by optimal choices",
    options = list(pageLength = 10, scrollX = TRUE),
    rownames = FALSE
  )
```

### Total Portfolio Effect

The [`hedged_portfolio()`](../reference/hedged_portfolio.html) function calculates total life expectancy gain from adopting all optimal lifestyle choices:

```{r}
portfolio <- hedged_portfolio()

# Summary by disease category
portfolio$by_category |>
  DT::datatable(
    caption = "Microlives gained by disease category",
    options = list(pageLength = 5, dom = "t"),
    rownames = FALSE
  )
```

```{r}
# Overall portfolio impact
portfolio$portfolio_summary |>
  DT::datatable(
    caption = "Total hedged portfolio effect",
    options = list(pageLength = 6, dom = "t"),
    rownames = FALSE
  )
```

**Interpretation:** A fully "hedged" individual (non-smoker, regular exercise, healthy diet, vaccinated, etc.) can expect to gain approximately `r portfolio$portfolio_summary$value[6]` days of life expectancy over 40 years compared to an "unhedged" baseline.

## 8. Conclusion

The `micromort` package helps translate abstract probabilities into concrete units for better decision-making. By comparing acute risks (micromorts), chronic risks (microlives), and quality-of-life metrics (QALYs, DALYs), individuals and policymakers can make more informed choices about risk trade-offs.

The new conditional risk functions enable:

*   **Cancer risk assessment:** Compare baseline risk to family history scenarios
*   **Vaccination value:** Quantify micromorts avoided through vaccination schedules
*   **Lifestyle optimization:** Calculate total life expectancy gain from adopting optimal "hedged" behaviors
