---
title: "Introduction to Micromorts and Risk Visualization"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to Micromorts and Risk Visualization}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(micromort)
library(ggplot2)
library(dplyr)
```

This vignette introduces the **micromort** package, which provides tools for understanding and visualizing risks.

## 1. Micromorts (Acute Risk)

A **micromort** is a unit of risk representing a one-in-a-million chance of death. It is used to measure acute risks—risks that can kill you immediately (e.g., skydiving, driving).

```{r}
# 1 in 10,000 chance of death = 100 micromorts
prob <- 1/10000
as_micromort(prob)

# Compare common risks
risks <- common_risks()
print(risks)
```

### Visualizing Risks

Using `plot_risks()`, we can see the relative magnitude of different activities on a logarithmic scale.

```{r, fig.width=7, fig.height=5}
plot_risks()
```

## 2. Microlives (Chronic Risk)

While micromorts measure sudden death, **microlives** measure the impact of chronic habits on your life expectancy. A microlife represents a 30-minute change in life expectancy.

Common chronic risks:
*   Smoking 1 cigarette: -1 microlife (approx 1 micromort equivalent risk)
*   Being 5kg overweight: -1 microlife per day
*   First 20 mins moderate exercise: +2 microlives

```{r}
# If smoking 20 cigarettes a day costs 1 microlife each (approx 30 mins)
daily_loss_minutes <- 20 * 30
daily_loss_microlives <- as_microlife(daily_loss_minutes)
print(daily_loss_microlives) # 20 microlives lost per day
```

## 3. Value of Statistical Life (VSL)

The **Value of a Statistical Life (VSL)** is the monetary value used to justify safety spending. It is NOT the value of an individual life, but the aggregate willingness to pay for small risk reductions.

Example: If a safety feature costs $50 and saves 1 life in 100,000 people (10 micromorts), is it worth it?
Cost per micromort saved = $50 / 10 = $5.
If VSL = $10M, then 1 micromort = $10. Since $5 < $10, it is cost-effective.

```{r}
# Standard VSL of $10M implies $10 per micromort
value_of_micromort(vsl = 10000000)

# Higher VSL implies higher safety spending
value_of_micromort(vsl = 15000000)
```

## 4. Loss of Life Expectancy (LLE)

**LLE** estimates the average time lost from a lifespan due to a specific risk. For a 1-in-a-million risk (1 micromort), the LLE is tiny.

```{r}
# Loss of life expectancy from 1 micromort (assuming 40 years remaining)
lle_minutes <- lle(prob = 1/1e6, life_expectancy = 40)
print(lle_minutes)
# Result is in minutes. 1 micromort ~ 21 minutes lost?
# No, 1 micromort = 1e-6 * 40 years = 40e-6 years = ~21 minutes
# Wait, check calculation:
# 40 years * 365.25 days * 24 hours * 60 minutes = ~21 million minutes
# 1e-6 * 21 million = ~21 minutes
```

## 5. QALY / DALY (Brief Overview)

*   **QALY (Quality-Adjusted Life Year):** Measures years of life adjusted for quality. 1 QALY = 1 year of perfect health. Used to assess medical interventions.
*   **DALY (Disability-Adjusted Life Year):** Measures years of life lost due to premature death + years lived with disability. Used to assess disease burden.

These metrics go beyond simple mortality risk to capture morbidity and quality of life.

## Conclusion

The `micromort` package helps translate abstract probabilities into concrete units for better decision-making. By comparing acute risks (micromorts) and chronic risks (microlives), individuals and policymakers can make more informed choices.
