---
title: "Introduction to Micromorts and Risk Visualization"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to Micromorts and Risk Visualization}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(micromort)
library(ggplot2)
library(dplyr)
library(DT)
```

This vignette introduces the **micromort** package, which provides tools for understanding and visualizing risks.

## 1. Micromorts (Acute Risk)

A **micromort** is a unit of risk representing a one-in-a-million chance of death. It is used to measure acute risks—risks that can kill you immediately (e.g., skydiving, driving).

```{r}
# 1 in 10,000 chance of death = 100 micromorts
prob <- 1/10000
as_micromort(prob)

# Compare common risks (sortable table - click column headers)
risks <- common_risks()
DT::datatable(
  risks,
  caption = "Common risks in micromorts (click column headers to sort, use search box to filter)",
  filter = "top",
  options = list(pageLength = 10, scrollX = TRUE),
  rownames = FALSE
)
```

### Visualizing Risks

Using [`plot_risks()`](../reference/plot_risks.html), we can see the relative magnitude of different activities on a logarithmic scale. The plot is split into COVID-19 and Other risks to make comparisons easier:
```{r, fig.width=8, fig.height=14}
# Default: faceted by COVID-19 vs Other
plot_risks()
```

#### Interactive Version

For interactive exploration with hover details and category filtering, use [`plot_risks_interactive()`](../reference/plot_risks_interactive.html):

```{r}
# plotly version with dropdown filter and hover details
plot_risks_interactive()
```

## 2. Microlives (Chronic Risk)

While micromorts measure sudden death, **microlives** measure the impact of chronic habits on your life expectancy. A microlife represents a 30-minute change in life expectancy.

Common chronic risks:
*   Smoking 1 cigarette: -1 microlife (approx 1 micromort equivalent risk)
*   Being 5kg overweight: -1 microlife per day
*   First 20 mins moderate exercise: +2 microlives

```{r}
# as_microlife() converts minutes of life expectancy to microlives
# where 1 microlife = 30 minutes of life expectancy

# Example 1: A heavy smoker (20 cigarettes/day)
# Each cigarette costs ~15-30 minutes of life expectancy
# Total daily cost: 20 cigarettes × 30 minutes = 600 minutes
as_microlife(600)  # = 20 microlives lost per day

# Example 2: Moderate exercise (20 mins) gains ~60 mins life expectancy
as_microlife(60)   # = 2 microlives gained

# Example 3: Being 5kg overweight costs ~30 mins per day
as_microlife(30)   # = 1 microlife lost per day

# Compare: How much exercise offsets smoking 2 cigarettes?
# Smoking 2 cigs = 60 mins lost = 2 microlives
# Exercise 20 mins = 60 mins gained = 2 microlives
# They roughly cancel out!
```

## 3. Relationship Between Micromorts and Microlives

### Theoretical Conversion

Micromorts (acute risk) and microlives (chronic attrition) measure different phenomena, but they can be approximately converted:

**Key relationship:** 1 micromort ≈ 0.7 microlives (assuming 40 years remaining life expectancy)

**Mathematical derivation:**

```{r}
# 1 micromort = 1-in-a-million death probability
# Expected life lost = probability × remaining life expectancy
remaining_years <- 40
prob_death <- 1e-6  # 1 micromort

# Life lost in years
life_lost_years <- prob_death * remaining_years
cat("Life lost (years):", life_lost_years, "\n")

# Convert to minutes
life_lost_minutes <- life_lost_years * 365.25 * 24 * 60
cat("Life lost (minutes):", round(life_lost_minutes, 1), "\n")

# 1 microlife = 30 minutes
microlives_equivalent <- life_lost_minutes / 30
cat("Microlives equivalent:", round(microlives_equivalent, 2), "\n")
```

This is why the `microlives` column in `common_risks()` uses the conversion `microlives = micromorts × 0.7`.

### Statistical Properties

| Metric | Best For | Statistical Properties |
|--------|----------|----------------------|
| **Micromorts** | Rare acute events (traffic deaths, skydiving) | Poisson-like; large samples needed; stable rates |
| **Microlives** | Chronic exposures (smoking, diet) | Cumulative effects; epidemiological cohort studies |

### When to Use Each

*   **Use micromorts** when comparing discrete events with immediate risk (surgery, travel, sports)
*   **Use microlives** when assessing lifestyle changes over time (quitting smoking, weight loss)
*   **Convert between them** when making policy decisions that involve both acute and chronic risks

## 4. Value of Statistical Life (VSL)

The **Value of a Statistical Life (VSL)** is the monetary value used to justify safety spending. It is NOT the value of an individual life, but the aggregate willingness to pay for small risk reductions.

Example: If a safety feature costs $50 and saves 1 life in 100,000 people (10 micromorts), is it worth it?
Cost per micromort saved = $50 / 10 = $5.
If VSL = $10M, then 1 micromort = $10. Since $5 < $10, it is cost-effective.

```{r}
# Standard VSL of $10M implies $10 per micromort
value_of_micromort(vsl = 10000000)

# Higher VSL implies higher safety spending
value_of_micromort(vsl = 15000000)
```

## 5. Loss of Life Expectancy (LLE)

**LLE** estimates the average time lost from a lifespan due to a specific risk. For a 1-in-a-million risk (1 micromort), the LLE is tiny.

```{r}
# Loss of life expectancy from 1 micromort (assuming 40 years remaining)
lle_minutes <- lle(prob = 1/1e6, life_expectancy = 40)
print(lle_minutes)
# Result is in minutes. 1 micromort ~ 21 minutes lost?
# No, 1 micromort = 1e-6 * 40 years = 40e-6 years = ~21 minutes
# Wait, check calculation:
# 40 years * 365.25 days * 24 hours * 60 minutes = ~21 million minutes
# 1e-6 * 21 million = ~21 minutes
```

## 6. Complementary Metrics: QALY, DALY, and Morbidity

Micromorts and microlives focus on mortality. But many conditions (like the common cold) cause significant quality of life loss without being fatal. Complementary metrics capture this morbidity burden.

### QALY (Quality-Adjusted Life Year)

Measures years of life adjusted for quality. **1 QALY = 1 year of perfect health.**

*   Health states are weighted 0 (death) to 1 (perfect health)
*   A year with chronic pain at 0.7 quality = 0.7 QALYs
*   Used to assess cost-effectiveness of medical interventions (e.g., £20,000-30,000 per QALY threshold in UK)

### DALY (Disability-Adjusted Life Years)

Measures disease burden as the sum of:

*   **YLL (Years of Life Lost):** From premature mortality
*   **YLD (Years Lived with Disability):** From morbidity, weighted by disability severity

**Formula:** `DALY = YLL + YLD`

For fatal diseases like COVID-19, YLL dominates. For non-fatal conditions like the common cold, YLD dominates.

### Capturing Non-Fatal Burden: The "Microburden" Concept

The common cold doesn't kill many people, but millions of work/school days are lost annually. How do we capture this in a microlife-like metric?

**Proposed "Microburden":** Population-level quality of life lost per million person-days.

```{r}
# Example: Common cold burden calculation
# Disability weight for common cold: ~0.006 (WHO GBD)
# Average duration: 5-7 days
# Episodes per person per year: ~2-3

cold_disability_weight <- 0.006
cold_duration_days <- 6
episodes_per_year <- 2.5

# QALD (Quality-Adjusted Life Days) lost per episode
qald_per_episode <- cold_disability_weight * cold_duration_days

# Annual burden per person (in days of perfect health equivalent)
annual_burden_days <- qald_per_episode * episodes_per_year
cat("Annual quality burden (QALD/person):", round(annual_burden_days, 4), "\n")

# For 1 million people
population <- 1e6
total_qald_lost <- annual_burden_days * population
cat("Population burden (QALD/million):", format(total_qald_lost, big.mark = ","), "\n")
```

### Comparing Metrics

| Metric | Focus | Unit | Best For |
|--------|-------|------|----------|
| Micromort | Acute mortality | 1-in-million death risk | Comparing immediate dangers |
| Microlife | Chronic mortality | 30 min life expectancy | Lifestyle interventions |
| QALY | Quality + quantity | 1 year perfect health | Healthcare economics |
| DALY | Disease burden | Years lost/disabled | Global health priorities |
| QALD | Daily morbidity | 1 day perfect health | Common illnesses |

### References

*   WHO Global Burden of Disease: [ghdx.healthdata.org](https://ghdx.healthdata.org/)
*   Spiegelhalter D (2012). BMJ 2012;345:e8223. doi:10.1136/bmj.e8223

## 7. Conditional Risks: Cancer, Vaccination, and Risk Hedging

### Cancer Risks by Type and Sex

The [`cancer_risks()`](../reference/cancer_risks.html) function provides mortality data stratified by cancer type, sex, and age group:

```{r}
# Top 3 cancers by mortality for each sex
cancer_risks() |>
  dplyr::filter(age_group == "All ages", sex != "Both") |>
  dplyr::group_by(sex) |>
  dplyr::slice_min(rank_by_sex, n = 3) |>
  dplyr::select(cancer_type, sex, deaths_per_100k, micromorts_per_year, family_history_rr) |>
  DT::datatable(
    caption = "Top 3 cancers by mortality rate (click to sort)",
    options = list(pageLength = 6, dom = "t"),
    rownames = FALSE
  )
```

**Family history impact:** The `family_history_rr` column shows relative risk increase with a first-degree relative's diagnosis. For example, prostate cancer risk increases 2.5× with family history.

```{r}
# Compare risk with vs without family history (male, age 50-64)
cancer_risks() |>
  dplyr::filter(sex == "Male", age_group == "50-64") |>
  dplyr::select(cancer_type, micromorts_per_year, micromorts_with_family_history) |>
  dplyr::mutate(
    increase_mm = micromorts_with_family_history - micromorts_per_year
  ) |>
  DT::datatable(
    caption = "Cancer risk with family history (Male, 50-64 years)",
    options = list(pageLength = 5, dom = "t"),
    rownames = FALSE
  )
```

### Vaccination Risk Reduction

The [`vaccination_risks()`](../reference/vaccination_risks.html) function quantifies micromorts avoided through vaccination:

```{r}
# Childhood vaccination impact by country
vaccination_risks() |>
  dplyr::filter(age_group == "0-5", grepl("Complete", vaccine_schedule)) |>
  dplyr::select(country, mortality_reduction_pct, micromorts_avoided_per_year, annual_life_days_gained) |>
  DT::datatable(
    caption = "Complete childhood vaccination schedule impact",
    options = list(pageLength = 5, dom = "t"),
    rownames = FALSE
  )
```

```{r}
# Adult vaccination benefits (age 65+)
vaccination_risks() |>
  dplyr::filter(age_group == "65+", country == "US") |>
  dplyr::select(vaccine_schedule, micromorts_avoided_per_year, microlives_gained_per_day) |>
  DT::datatable(
    caption = "Adult vaccination benefits (US, age 65+)",
    options = list(pageLength = 5, dom = "t"),
    rownames = FALSE
  )
```

### Hedged vs Unhedged: Optimal Lifestyle Comparison

The [`conditional_risk()`](../reference/conditional_risk.html) function compares risk factors between optimal ("hedged") and suboptimal ("unhedged") states:

```{r}
# Cardiovascular risk hedging
conditional_risk("cardiovascular") |>
  dplyr::select(risk_factor, unhedged_state, hedged_state, microlives_gained, annual_days_gained) |>
  DT::datatable(
    caption = "Cardiovascular hedging: microlives gained by optimal choices",
    options = list(pageLength = 10, scrollX = TRUE),
    rownames = FALSE
  )
```

### Total Portfolio Effect

The [`hedged_portfolio()`](../reference/hedged_portfolio.html) function calculates total life expectancy gain from adopting all optimal lifestyle choices:

```{r}
portfolio <- hedged_portfolio()

# Summary by disease category
portfolio$by_category |>
  DT::datatable(
    caption = "Microlives gained by disease category",
    options = list(pageLength = 5, dom = "t"),
    rownames = FALSE
  )
```

```{r}
# Overall portfolio impact
portfolio$portfolio_summary |>
  DT::datatable(
    caption = "Total hedged portfolio effect",
    options = list(pageLength = 6, dom = "t"),
    rownames = FALSE
  )
```

**Interpretation:** A fully "hedged" individual (non-smoker, regular exercise, healthy diet, vaccinated, etc.) can expect to gain approximately `r portfolio$portfolio_summary$value[6]` days of life expectancy over 40 years compared to an "unhedged" baseline.

## 8. Conclusion

The `micromort` package helps translate abstract probabilities into concrete units for better decision-making. By comparing acute risks (micromorts), chronic risks (microlives), and quality-of-life metrics (QALYs, DALYs), individuals and policymakers can make more informed choices about risk trade-offs.

The new conditional risk functions enable:

*   **Cancer risk assessment:** Compare baseline risk to family history scenarios
*   **Vaccination value:** Quantify micromorts avoided through vaccination schedules
*   **Lifestyle optimization:** Calculate total life expectancy gain from adopting optimal "hedged" behaviors
