---
title: "Introduction to Micromorts and Risk Visualization"
format:
  html:
    code-fold: true
    code-summary: "Show code"
vignette: >
  %\VignetteIndexEntry{Introduction to Micromorts and Risk Visualization}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

if (requireNamespace("micromort", quietly = TRUE)) {
  library(micromort)
} else {
  pkgload::load_all(quiet = TRUE)
}

library(targets)
library(DT)

# Safe tar_read with graceful fallback (finds project root for pkgdown builds)
safe_tar_read <- function(name) {
  store <- Sys.getenv("TARGETS_STORE", unset = NA)
  if (is.na(store)) {
    store <- tryCatch(
      file.path(rprojroot::find_root(rprojroot::is_r_package), "_targets"),
      error = function(e) "_targets"
    )
  }
  tryCatch(
    targets::tar_read_raw(name, store = store),
    error = function(e) {
      message("Target '", name, "' not found. Run tar_make() first.")
      NULL
    }
  )
}
```

This vignette introduces the **micromort** package, which provides tools for understanding and visualizing risks.

## 1. Micromorts (Acute Risk)

A **micromort** is a unit of risk representing a one-in-a-million chance of death. More precisely, it's a **microprobability** of death — a one-in-a-million chance of a specific event (death) occurring.

### Definition

| Term | Definition | Example |
|------|------------|---------|
| **Microprobability** | 1-in-a-million chance of any event | 1 micromort = microprobability of death |
| **Micromort** | 1-in-a-million chance of death, per event | Skydiving: 8 micromorts per jump |

### Comparing Risks: Period Matters!

**CRITICAL:** When comparing micromort values, ensure the **period is the same**. For example:

| Activity | Micromorts | Period | Comparable? |
|----------|------------|--------|-------------|
| Scuba diving | 5 | per dive | Per-event |
| Scuba diving | 164 | per year | Per-year (assumes ~33 dives/year) |
| Skydiving | 8 | per jump | Per-event |

The "per year" figure (164) conflates frequency with risk-per-event. A diver doing 5 dives/year vs 50 dives/year faces very different annual risk.

### Conditional Risks

Many activities have **selection effects**. For example:

*   **Marathon running (7 micromorts):** Runners are self-selected for fitness. This low figure reflects the health of participants, not the risk to an average person attempting a marathon.
*   **Motorcycle riding (10 micromorts/60 miles):** Experienced riders face lower risk than novices, but the quoted figure is an average.

These figures answer: "Given that someone completed this activity, what was their death risk?" Not: "What would happen if a random person attempted this?"

### Converting Probabilities to Micromorts

```{r introduction-chunk-1}
# 1 in 10,000 chance of death = 100 micromorts
as_micromort(1/10000)
```

### Common Risks Table

```{r introduction-chunk-2}
#| echo: false
safe_tar_read("vig_intro_common_risks") |>
  DT::datatable(
    caption = "Common risks in micromorts (click column headers to sort, use search box to filter)",
    filter = "top",
    options = list(pageLength = 10, scrollX = TRUE),
    rownames = FALSE
  )
```

### Visualizing Risks

Using [`plot_risks()`](../reference/plot_risks.html), we can see the relative magnitude of different activities on a logarithmic scale. The plot is split into COVID-19 and Other risks to make comparisons easier:

```{r introduction-chunk-3, fig.width=10, fig.height=14, out.extra='style="background-color: #1a1a1a; padding: 10px; border-radius: 4px;"'}
#| echo: false
#| fig-cap: "Logarithmic risk ladder showing micromorts per activity, split by COVID-19 and other categories. Activities range from a chest X-ray (0.1 micromort) to BASE jumping (430 micromorts)."
#| fig-alt: "Horizontal bar chart on a log scale showing micromorts for ~40 activities. COVID-19 risks (vaccination, infection by age) are grouped separately from other risks (transport, medical, recreational). Bars are coloured by category."
print(safe_tar_read("vig_intro_risk_plot"))
```

#### Interactive Version

For interactive exploration with hover details and category filtering, use [`plot_risks_interactive()`](../reference/plot_risks_interactive.html):

```{r introduction-chunk-4}
#| echo: false
#| fig-cap: "Interactive version of the risk ladder with hover details and category filtering."
#| fig-alt: "Interactive plotly bar chart of micromorts per activity with tooltips showing exact values, microlives, and source period."
safe_tar_read("vig_intro_risk_plot_interactive")
```

## 2. Microlives (Chronic Risk)

While micromorts measure sudden death, **microlives** measure the impact of chronic habits on your life expectancy. A microlife represents a 30-minute change in life expectancy **per day of exposure**.

### Living at Different Speeds

A useful way to think about microlives: we all "use up" **48 microlives per day** just by living (24 hours = 48 × 30 minutes). Unhealthy habits accelerate this consumption, while healthy habits slow it down.

A smoker who smokes 20 cigarettes per day uses up an additional 10 microlives, which can be interpreted as **rushing towards death at 29 hours per day** instead of 24. Conversely, someone with excellent lifestyle habits might effectively live at only 22 hours per day.

**Healthcare bonus:** Modern healthcare and healthier lifestyles give us a "payback" of approximately **12 microlives per day** — our expected death is moving away from us even as we age.

### Common Chronic Risks

| Factor | Microlives/day | Interpretation |
|--------|----------------|----------------|
| Smoking 1 cigarette | -1 | Lose 30 min life expectancy |
| Being 5kg overweight | -1 | Lose 30 min/day |
| 20 min moderate exercise | +2 | Gain 60 min life expectancy |
| 2+ hours TV daily | -1 | Sedentary behavior |
| 5+ servings fruit/veg | +2 | Healthy diet |

### Converting Life Expectancy to Microlives

```{r introduction-chunk-5}
# as_microlife() converts minutes of life expectancy change to microlives
# Unit: 1 microlife = 30 minutes of life expectancy change PER DAY
# Sign: negative = loss, positive = gain

# Heavy smoker: 20 cigarettes/day, each costs ~30 mins
as_microlife(-20 * 30)  # = -20 microlives/day (life lost)

# Moderate exercise: 20 mins → ~60 mins life gained
as_microlife(60)        # = +2 microlives/day (life gained)

# Being 5kg overweight costs 30 mins per day
as_microlife(-30)       # = -1 microlife/day (life lost)
```

## 3. Relationship Between Micromorts and Microlives

### Theoretical Conversion

Micromorts (acute, per-event risk) and microlives (chronic, per-day attrition) measure different phenomena, but can be approximately converted using expected value theory.

**Key relationship:** 1 micromort ≈ 0.7 microlives

**Assumptions for this conversion:**
1. Remaining life expectancy = 40 years (use `lle(prob, life_expectancy = ...)` to [adjust for any age](rest_api.html#age-based-hazard-rates); see also `daily_hazard_rate()`)
2. Death occurs immediately upon the event (worst case)
3. Linear approximation (valid for small probabilities)

**Mathematical derivation:**

- 1 micromort = 1/1,000,000 probability of death
- Expected life lost = probability × remaining life
- = 1e-6 × 40 years = 4e-5 years
- = 4e-5 × 525,960 minutes/year ≈ 21 minutes
- 1 microlife = 30 minutes, so 21 minutes ≈ 0.7 microlives

**Why `common_risks()` uses microlives = micromorts × 0.7:**

The conversion allows comparing a single risky event (like one skydive at 8 micromorts) to chronic daily habits (like smoking at -10 microlives/day). However, **the approximation breaks down when:**

1. **Remaining life expectancy differs** from 40 years (a 20-year-old loses more per micromort than an 80-year-old)
2. **The risk is not immediate death** (injuries, disabilities are not captured)
3. **Repeated exposures** compound non-linearly

### Unit Definitions (Summary)

| Metric | Unit Definition | Scope | Sign |
|--------|-----------------|-------|------|
| **Micromort** | 1-in-a-million probability of death | Per discrete event (1 surgery, 1 flight) | Always ≥ 0 |
| **Microlife** | 30 minutes of life expectancy change per day | Per day of exposure/habit | + = gain, − = loss |

### When to Use Each

*   **Use micromorts** for discrete, short-duration events with binary outcomes (death or survival): surgery, skydiving, a single car trip
*   **Use microlives** for chronic, daily habits that accumulate over a lifetime: smoking, exercise, diet
*   **Convert between them** for policy decisions, but state your assumptions (age, remaining life expectancy)

## 4. Value of Statistical Life (VSL)

The **Value of a Statistical Life (VSL)** is the monetary value used to justify safety spending. It is NOT the value of an individual life, but the aggregate willingness to pay for small risk reductions.

### US Valuation

Example: If a safety feature costs $50 and saves 1 life in 100,000 people (10 micromorts), is it worth it?
Cost per micromort saved = $50 / 10 = $5.
If VSL = $10M, then 1 micromort = $10. Since $5 < $10, it is cost-effective.

```{r introduction-chunk-6}
# Standard US VSL of $10M implies $10 per micromort
value_of_micromort(vsl = 10000000)
```

### UK Valuation: Micromorts ≈ Microlives

Interestingly, two UK government agencies arrive at similar valuations for micromorts and microlives:

| Agency | Metric | Value | Per Unit |
|--------|--------|-------|----------|
| **NICE** (NHS) | 1 QALY | ~£30,000 | **£1.70 per microlife** |
| **Dept of Transport** | Value of Statistical Life | £1,600,000 | **£1.60 per micromort** |

This near-equivalence (£1.60 ≈ £1.70) provides empirical support for the theoretical conversion: **1 micromort ≈ 1 microlife** in policy terms.

```{r introduction-chunk-7}
# UK Department of Transport VSL: £1.6M → £1.60 per micromort
value_of_micromort(vsl = 1600000)
```

This consistency suggests that policy decisions affecting acute risks (transport safety) and chronic risks (healthcare interventions) can be compared on a common scale.

## 5. Loss of Life Expectancy (LLE)

**LLE** estimates the average time lost from a lifespan due to a specific risk. For a 1-in-a-million risk (1 micromort), the LLE is approximately 21 minutes (assuming 40 years remaining life).

```{r introduction-chunk-8}
# Loss of life expectancy from 1 micromort (assuming 40 years remaining)
lle(prob = 1/1e6, life_expectancy = 40)
```

## 6. Complementary Metrics: QALY, DALY, and Morbidity

Micromorts and microlives focus on mortality. But many conditions (like the common cold) cause significant quality of life loss without being fatal. Complementary metrics capture this morbidity burden.

### QALY (Quality-Adjusted Life Year)

Measures years of life adjusted for quality. **1 QALY = 1 year of perfect health.**

*   Health states are weighted 0 (death) to 1 (perfect health)
*   A year with chronic pain at 0.7 quality = 0.7 QALYs
*   Used to assess cost-effectiveness of medical interventions (e.g., £20,000-30,000 per QALY threshold in UK)

### DALY (Disability-Adjusted Life Years)

Measures disease burden as the sum of two components:

*   **DALY = YLL + YLD**, where:
    *   **YLL (Years of Life Lost):** From premature mortality
    *   **YLD (Years Lived with Disability):** From morbidity, weighted by disability severity

For fatal diseases like COVID-19, YLL dominates. For non-fatal conditions like the common cold, YLD dominates.

### Comparing Metrics

| Metric | Unit Definition | Scope | Sign | Best For |
|--------|-----------------|-------|------|----------|
| **Micromort** | 1/1,000,000 death probability | Per discrete event (surgery, flight, climb) | ≥ 0 (probability) | Comparing single risky activities |
| **Microlife** | 30 min life expectancy change per day | Per day of chronic exposure | + gain / − loss | Daily lifestyle interventions |
| **QALY** | 1 year at perfect health (quality=1.0) | Per treatment/intervention | ≥ 0 | Cost-effectiveness in healthcare |
| **DALY** | 1 year lost to disease (YLL + YLD) | Per condition/population | ≥ 0 (burden) | Global health prioritization |
| **QALD** | 1 day at perfect health | Per illness episode | ≥ 0 | Short-term morbidity (colds, flu) |

### References

*   Spiegelhalter D (2012). "Using speed of ageing and 'microlives'." BMJ 2012;345:e8223. [doi:10.1136/bmj.e8223](https://doi.org/10.1136/bmj.e8223)
*   Spiegelhalter D (2012). "Understanding uncertainty: Microlives." Plus Magazine. [plus.maths.org](https://plus.maths.org/content/understanding-uncertainty-microlives)
*   WHO Global Burden of Disease: [ghdx.healthdata.org](https://ghdx.healthdata.org/)
*   NICE Methods Guide: [nice.org.uk](https://www.nice.org.uk/process/pmg9/chapter/the-reference-case)

## 7. Conditional Risks: Cancer, Vaccination, and Risk Hedging

### Cancer Risks by Type and Sex

The [`cancer_risks()`](../reference/cancer_risks.html) function provides mortality data stratified by cancer type, sex, and age group:

```{r introduction-chunk-9}
#| echo: false
safe_tar_read("vig_intro_cancer_top3") |>
  DT::datatable(
    caption = "Top 3 cancers by mortality rate (click to sort)",
    options = list(pageLength = 6, dom = "t"),
    rownames = FALSE
  )
```

**Family history impact:** The `family_history_rr` column shows relative risk increase with a first-degree relative's diagnosis. For example, prostate cancer risk increases 2.5× with family history.

```{r introduction-chunk-10}
#| echo: false
safe_tar_read("vig_intro_cancer_family_history") |>
  DT::datatable(
    caption = "Cancer risk with family history (Male, 50-64 years)",
    options = list(pageLength = 5, dom = "t"),
    rownames = FALSE
  )
```

### Vaccination Risk Reduction

The [`vaccination_risks()`](../reference/vaccination_risks.html) function quantifies micromorts avoided through vaccination:

```{r introduction-chunk-11}
#| echo: false
safe_tar_read("vig_intro_vaccination_childhood") |>
  DT::datatable(
    caption = "Complete childhood vaccination schedule impact",
    options = list(pageLength = 5, dom = "t"),
    rownames = FALSE
  )
```

```{r introduction-chunk-12}
#| echo: false
safe_tar_read("vig_intro_vaccination_adult") |>
  DT::datatable(
    caption = "Adult vaccination benefits (US, age 65+)",
    options = list(pageLength = 5, dom = "t"),
    rownames = FALSE
  )
```

### Hedged vs Unhedged: Optimal Lifestyle Comparison

The [`conditional_risk()`](../reference/conditional_risk.html) function compares risk factors between optimal ("hedged") and suboptimal ("unhedged") states:

```{r introduction-chunk-13}
#| echo: false
safe_tar_read("vig_intro_cardiovascular_risk") |>
  DT::datatable(
    caption = "Cardiovascular hedging: microlives gained by optimal choices",
    options = list(pageLength = 10, scrollX = TRUE),
    rownames = FALSE
  )
```

### Total Portfolio Effect

The [`hedged_portfolio()`](../reference/hedged_portfolio.html) function calculates total life expectancy gain from adopting all optimal lifestyle choices:

```{r introduction-chunk-14}
#| echo: false
portfolio <- safe_tar_read("vig_intro_hedged_portfolio")
if (!is.null(portfolio)) {
  portfolio$by_category |>
    DT::datatable(
      caption = "Microlives gained by disease category",
      options = list(pageLength = 5, dom = "t"),
      rownames = FALSE
    )
}
```

```{r introduction-chunk-15}
#| echo: false
if (!is.null(portfolio)) {
  portfolio$portfolio_summary |>
    DT::datatable(
      caption = "Total hedged portfolio effect",
      options = list(pageLength = 6, dom = "t"),
      rownames = FALSE
    )
}
```

**Interpretation:** A fully "hedged" individual (non-smoker, regular exercise, healthy diet, vaccinated, etc.) can expect to gain significant additional life expectancy compared to an "unhedged" baseline.

### Conditional Acute Risks: Age Changes Everything

Conditional risk analysis isn't limited to chronic lifestyle factors. Some acute risks show extreme **age-conditioning** that makes population averages misleading.

**Example: Falling out of bed** kills ~450 Americans per year ([CPSC](https://www.cpsc.gov/Newsroom/News-Releases/2022/Older-Americans-Are-More-Likely-to-Suffer-Fatalities-from-Falls-and-Fire-CPSC-Report-Highlights-Hidden-Hazards-Around-the-Home)). The population average is 1.36 micromorts/year — but the [CDC age-stratified data](https://www.cdc.gov/nchs/products/databriefs/db532.htm) reveals a **2,500-fold** difference across age groups:

| Age group | Sex | Fall deaths per 100,000/year | Micromorts per night of sleep |
|-----------|-----|------------------------------|------------------------------|
| Under 65 | Both | ~0.4 | **0.004** |
| 65-74 | Male | 24.7 | **0.68** |
| 65-74 | Female | 14.2 | **0.39** |
| 85+ | Male | 373.3 | **10.2** |
| 85+ | Female | 319.7 | **8.8** |

An 85-year-old man going to bed faces ~10 micromorts per night — comparable to riding a motorcycle 60 miles (10 micromorts/trip) or a single dose of ecstasy (13 micromorts/dose). For someone under 65, the same activity carries 0.004 micromorts per night — essentially zero.

This pattern recurs across many "exotic" risks:

*   **Bee/wasp stings** (72 deaths/year, US): Nearly all fatalities are among the ~1% of the population with venom allergy. Conditional on allergy: high risk per sting. Conditional on no allergy: near-zero.
*   **Cow trampling** (22 deaths/year, US): Population rate is 0.07 micromorts/year. For cattle farmers handling animals daily: ~7.5 micromorts/year — a 100-fold increase.
*   **Lightning strike** (28 deaths/year, US): Population rate is 0.08 micromorts/year. For outdoor agricultural workers: ~1.2 micromorts/year — a 15-fold increase.

The lesson: **always ask "conditional on what?"** before comparing micromort values across populations or activities. For a deeper treatment of how confounding variables distort risk data, including Simpson's paradox and stratification strategies, see the [Confounding Variables](confounding.html) vignette.

## 8. Data Quality: The Denominator Problem

Not every widely-cited risk statistic belongs in a curated dataset. Many "exotic" risk figures circulate in media and popular science without the rigorous **exposure denominator** needed for meaningful micromort calculations. This section explains why certain risks were excluded and what to watch for when interpreting risk data.

### 8.1 The Three Denominator Failures

A micromort value is only meaningful when paired with a clear **period** — the unit of exposure. We require: *Deaths ÷ Exposures = Probability*. Three common failures make this impossible:

**1. Unknown denominator (no exposure count)**

| Risk | Annual deaths (US) | What we don't know |
|------|-------------------|--------------------|
| Bee/wasp stings | 72 | How many sting *exposures* per year? Is this per sting, per outdoor hour, per year of beekeeping? |
| Dog mauling | 43-127 | Per dog interaction? Per year of dog ownership? Per year of existing near dogs? |
| Fire ant stings | ~30 | Per sting event? Per year in fire ant territory? |

The 72 bee sting deaths per year ([CDC MMWR](https://www.cdc.gov/mmwr/volumes/68/wr/mm6829a5.htm)) is a solid numerator from the National Vital Statistics System. But without knowing the total number of sting exposures, we cannot calculate a per-sting micromort value. The best we can do is a per-year population rate (0.22 micromorts/year for any US resident), which conflates people who never encounter bees with beekeepers stung weekly.

**2. Misleading population average (hides conditional risk)**

| Risk | Population average | Conditional reality |
|------|-------------------|---------------------|
| Falling out of bed | 1.36 micromorts/year (US) | Under-65: ~0.004/night; Age 85+ male: **10.2/night** |
| Cow trampling | 0.07 micromorts/year (US) | General public: ~0; Cattle farmer: ~7.5/year |
| Lightning strike | 0.08 micromorts/year (US) | Indoor worker: ~0.02; Outdoor agricultural worker: ~1.2/year |

Falling out of bed kills ~450 Americans per year ([CPSC](https://www.cpsc.gov/Newsroom/News-Releases/2022/Older-Americans-Are-More-Likely-to-Suffer-Fatalities-from-Falls-and-Fire-CPSC-Report-Highlights-Hidden-Hazards-Around-the-Home)), but the CDC age-stratified data ([Data Brief 532](https://www.cdc.gov/nchs/products/databriefs/db532.htm)) reveals a **2,500-fold** difference: an 85-year-old man faces ~10 micromorts per night of sleep (comparable to riding a motorcycle 60 miles), while someone under 65 faces ~0.004. The population average of 1.36/year is technically correct but practically useless.

**3. Fabricated or untraceable numerator**

| Claim | Cited figure | Actual evidence |
|-------|-------------|-----------------|
| Coconut deaths | "150 per year worldwide" | Originated from a **travel insurance press release** (Club Direct, late 1990s), not from Barss (1984). The original [Barss paper](https://pubmed.ncbi.nlm.nih.gov/6502774/) documented only 2 deaths over 4 years at one Papua New Guinea hospital. ([Snopes: Unproven](https://www.snopes.com/fact-check/coconuts-kill-more-sharks/)) |
| Champagne cork deaths | "24 per year" | Traced to a 2008 *Daily Mail* article citing a tabloid book. No government vital statistics, hospital registry, or peer-reviewed study supports this figure. |
| Left-handed equipment deaths | "9 years shorter lifespan" | A statistical artefact from [Halpern & Coren (1991)](https://pubmed.ncbi.nlm.nih.gov/2006231/), debunked by multiple papers. The apparent age gap reflects historical suppression of left-handedness reporting, not excess mortality ([van der Hoeven et al., 2023](https://pmc.ncbi.nlm.nih.gov/articles/PMC10369838/)). |

### 8.2 Inclusion Criteria for This Package

Risks in `common_risks()` and `acute_risks()` meet these minimum standards:

1. **Traceable numerator**: Death count from a government agency (CDC, CPSC, WHO, NOAA) or peer-reviewed epidemiological study
2. **Defined denominator**: A meaningful exposure unit (per ride, per climb, per day, per trip) — not just "per year of being alive"
3. **Reproducible calculation**: The micromort derivation can be independently verified from the cited source
4. **Not a myth**: The figure has not been debunked by subsequent peer-reviewed work

Risks that fail criterion 2 but pass the others may still appear with `period = "per year"` as a background population rate, clearly labelled as such. See `risk_data_sources()` for full provenance of each entry. For examples of how confounding variables (age, occupation, allergy status) can change risk estimates by orders of magnitude, see the [Confounding Variables](confounding.html) vignette.

### 8.3 When "Per Year" Is Legitimate

Some risks are genuinely **background hazards** where the exposure is simply *being alive in a given geography*:

*   Lightning strike: 0.08 micromorts/year (US) — everyone is exposed to thunderstorms
*   Asteroid impact: 0.01 micromorts/year (global) — a theoretical actuarial average ([NASA](https://ntrs.nasa.gov/api/citations/20160013841/downloads/20160013841.pdf))
*   Living one day at age 75: 120 micromorts/day — the baseline hazard rate from all causes

These are valid "per year" figures because the denominator is the entire population and the exposure is unavoidable. The problem arises when a per-year population rate is presented for an activity that only a fraction of the population engages in.

## 9. Conclusion

The `micromort` package helps translate abstract probabilities into concrete units for better decision-making. By comparing acute risks (micromorts), chronic risks (microlives), and quality-of-life metrics (QALYs, DALYs), individuals and policymakers can make more informed choices about risk trade-offs.

The new conditional risk functions enable:

*   **Cancer risk assessment:** Compare baseline risk to family history scenarios
*   **Vaccination value:** Quantify micromorts avoided through vaccination schedules
*   **Lifestyle optimization:** Calculate total life expectancy gain from adopting optimal "hedged" behaviors

## Reproducibility

```{r introduction-session-info, eval=TRUE}
sessionInfo()
```
