---
title: "Palatable Units: The Spiegelhalter Philosophy"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Palatable Units: The Spiegelhalter Philosophy}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(micromort)
library(dplyr)
library(ggplot2)
```

**"Statistics are not just numbers; they are the way we make sense of the world."** — Sir David Spiegelhalter

This vignette outlines the philosophy of **"Palatable Units"** championed by David Spiegelhalter (Winton Professor for the Public Understanding of Risk, Cambridge). His core argument is that abstract probabilities (e.g., "0.00004% hazard ratio") are meaningless to most people. To demystify risk, we must translate these into concrete, relatable units.

## 1. The Core Philosophy: Compare Apples to Oranges

The goal of palatable units is to create a common currency for risk. This allows us to strip away the emotional "dread factor" from scary-sounding events and compare them rationally against mundane activities.

### The Standard Units

*   **Micromort:** 1-in-a-million chance of **acute** death (sudden event).
*   **Microlife:** 30 minutes of life expectancy lost/gained (chronic **attrition**).

## 2. Micromorts: Measuring "Stopping Living"

A micromort measures the immediate risk of an event stopping your life.

### Comparative Risk Table

The following table demonstrates how disparate activities can be compared using the micromort. Notice how "scary" events like anesthesia carry similar risks to mundane travel.

```{r}
risks <- tibble::tribble(
  ~Activity, ~Micromorts, ~Category, ~Context,
  "Living (age 30)", 1, "Baseline", "Daily background risk for a young adult",
  "Driving 230 miles", 1, "Travel", "Routine long-distance drive",
  "Walking 20 miles", 1, "Travel", "Pedestrian accident risk",
  "Scuba Diving", 5, "Sport", "Per dive (PADI certified)",
  "Skydiving", 7, "Sport", "Per jump (USPA stats)",
  "Motorcycling 60 miles", 10, "Travel", "High-risk transport",
  "General Anesthesia", 10, "Medical", "Per emergency operation",
  "Running a Marathon", 7, "Sport", "Cardiovascular stress per race"
) %>% 
  arrange(desc(Micromorts))

knitr::kable(risks, caption = "Risk comparison in Micromorts (ordered by decreasing risk)")
```

> **Insight:** Riding a motorcycle for just 60 miles carries the same acute death risk (~10 micromorts) as undergoing general anesthesia.

## 3. Microlives: Measuring "Speed of Aging"

While micromorts measure sudden death, **Microlives** measure the **rate of attrition** caused by lifestyle habits.

*   **Definition:** 1 Microlife = 30 minutes of life expectancy.
*   **Logic:** A young adult (~30 years old) has roughly 1 million half-hours left to live. A microlife is one-millionth of their remaining lifespan.
*   **Sign:**
    *   **-1 Microlife:** You lose 30 minutes (aging faster).
    *   **+1 Microlife:** You gain 30 minutes (aging slower).

### Daily Habits Table

```{r}
chronic_risks <- tibble::tribble(
  ~Habit, ~Impact_Microlives, ~Time_Effect,
  "Smoking 2 cigarettes", -1, "Lose 30 minutes",
  "Being 5kg overweight", -1, "Lose 30 minutes (per day)",
  "Second pint of beer", -1, "Lose 30 minutes",
  "First 20 mins exercise", +1, "Gain 30 minutes",
  "Eating portion of fruit", +1, "Gain 30 minutes"
)

knitr::kable(chronic_risks, caption = "Impact of daily habits in Microlives")
```

> **Clarification:** A value of **-1 Microlife** is a **loss** of life. It means you have "spent" 30 minutes of your future life expectancy today. If you smoke 20 cigarettes (1 pack) a day, you lose ~10 microlives (5 hours) every day.

## 4. Visualization: The Risk Ladder

Spiegelhalter advocates for a **Logarithmic Risk Ladder**. This visualization helps placing rare risks (like asteroid impacts or terrorism) in context with daily risks.

*   **Why Log Scale?** Because risks span vast orders of magnitude (1 in 10 to 1 in 10 million).
*   **Interpretation:** A "100% increase" in a very rare risk (e.g., eating bacon increasing bowel cancer risk) might look huge in headlines but is often negligible on the ladder compared to the baseline risk of driving.

```{r, fig.width=7, fig.height=5}
# Using the package's built-in plotting function which implements this ladder concept
plot_risks()
```

## 5. Recommended Tools

While David Spiegelhalter focuses on concepts rather than specific software, the following R packages align with his mission of clear risk communication:

*   **`riskCommunicator`:** Designed for public health to provide interpretable effect measures (risk differences, number needed to treat) rather than abstract regression coefficients.
*   **`ggplot2`:** The standard for creating custom visuals like Risk Ladders and icon arrays.
*   **`micromort` (this package):** Specifically built to implement the palatable units framework.

## References

1.  Spiegelhalter, D., & Blastland, M. (2013). *The Norm Chronicles: Stories and numbers about danger*. Profile Books.
2.  Spiegelhalter, D. (2019). *The Art of Statistics: Learning from Data*. Pelican.
