---
title: "Palatable Units: The Spiegelhalter Philosophy"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Palatable Units: The Spiegelhalter Philosophy}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(micromort)
library(dplyr)
library(ggplot2)
```

**"Statistics are not just numbers; they are the way we make sense of the world."** — Sir David Spiegelhalter

This vignette outlines the philosophy of **"Palatable Units"** championed by David Spiegelhalter (Winton Professor for the Public Understanding of Risk, Cambridge). His core argument is that abstract probabilities (e.g., "0.00004% hazard ratio") are meaningless to most people. To demystify risk, we must translate these into concrete, relatable units.

## 1. The Core Philosophy: Compare Apples to Oranges

The goal of palatable units is to create a common currency for risk. This allows us to strip away the emotional "dread factor" from scary-sounding events and compare them rationally against mundane activities.

### The Standard Units

*   **Micromort:** 1-in-a-million chance of **acute** death (sudden event).
*   **Microlife:** 30 minutes of life expectancy lost/gained (chronic **attrition**).

## 2. Micromorts: Measuring "Stopping Living" (Hazard)

A micromort measures **acute hazard**: the immediate probability of an event causing death.

*   **Normalization:** Risk is normalized **per event** (or per unit distance), independent of the event's duration.
*   **Time Horizon:** The "time" is the discrete event itself.
    *   **Skydiving:** The risk is ~7 micromorts *per jump*. Whether the freefall lasts 30 seconds or 60 seconds is secondary to the event of jumping.
    *   **Scuba Diving:** The risk is ~5 micromorts *per dive*. A 30-minute dive and a 45-minute dive are treated as single "dive events" in broad statistics, though technically longer exposure increases risk.
    *   **Anesthesia:** ~10 micromorts *per operation*.

### Comparative Risk Table

The following table demonstrates how disparate activities can be compared using the micromort.

```{r}
risks <- tibble::tribble(
  ~Activity, ~Micromorts, ~Category, ~Context,
  "Motorcycling 60 miles", 10, "Travel", "High-risk transport",
  "General Anesthesia", 10, "Medical", "Per emergency operation",
  "Skydiving", 7, "Sport", "Per jump (USPA stats)",
  "Running a Marathon", 7, "Sport", "Cardiovascular stress per race",
  "Scuba Diving", 5, "Sport", "Per dive (PADI certified)",
  "Living (age 30)", 1, "Baseline", "Daily background risk for a young adult",
  "Driving 230 miles", 1, "Travel", "Routine long-distance drive",
  "Walking 20 miles", 1, "Travel", "Pedestrian accident risk"
)

knitr::kable(risks, caption = "Risk comparison in Micromorts (ordered by decreasing risk)")
```

> **Comparison:** Riding a motorcycle for just 60 miles carries the same acute death risk (~10 micromorts) as undergoing general anesthesia.

## 3. Microlives: Measuring "Speed of Aging" (Attrition)

While micromorts measure sudden death (Hazard), **Microlives** measure **chronic attrition**: the rate at which you are "using up" your life expectancy.

*   **Definition:** 1 Microlife = 30 minutes of life expectancy.
*   **Normalization:** Risk is normalized **per day** of maintaining a habit.
*   **Unit of Attrition:** The "unit" is the expected lifespan. -1 Microlife means your expected lifespan has shrunk by 30 minutes.
*   **Time Horizon:** Continuous. If you smoke 20 cigarettes a day, you are losing 10 microlives (5 hours) *every single day*.

### Daily Habits Table

```{r}
chronic_risks <- tibble::tribble(
  ~Habit, ~Impact_Microlives, ~Time_Effect,
  "Smoking 2 cigarettes", -1, "Lose 30 minutes life expectancy",
  "Being 5kg overweight", -1, "Lose 30 minutes (per day of being overweight)",
  "Second pint of beer", -1, "Lose 30 minutes",
  "First 20 mins exercise", +1, "Gain 30 minutes life expectancy",
  "Eating portion of fruit", +1, "Gain 30 minutes"
)

knitr::kable(chronic_risks, caption = "Impact of daily habits in Microlives")
```

> **Clarification:** A value of **-1 Microlife** is a **loss** (attrition). It effectively means you are aging 30 minutes faster than normal. Over a year, a -1 daily deficit sums to ~180 hours (7.5 days) of lost life.

## 4. Visualization: The Risk Ladder

Spiegelhalter advocates for a **Logarithmic Risk Ladder**. This visualization helps placing rare risks (like asteroid impacts or terrorism) in context with daily risks.

*   **Why Log Scale?** Because risks span vast orders of magnitude (1 in 10 to 1 in 10 million).
*   **Interpretation:** A "100% increase" in a very rare risk (e.g., eating bacon increasing bowel cancer risk) might look huge in headlines but is often negligible on the ladder compared to the baseline risk of driving.

```{r, fig.width=7, fig.height=5}
# Using the package's built-in plotting function which implements this ladder concept
plot_risks()
```

## 5. Recommended Tools

While David Spiegelhalter focuses on concepts rather than specific software, the following R packages align with his mission of clear risk communication:

*   **`riskCommunicator`:** Designed for public health to provide interpretable effect measures (risk differences, number needed to treat) rather than abstract regression coefficients.
*   **`ggplot2`:** The standard for creating custom visuals like Risk Ladders and icon arrays.
*   **`micromort` (this package):** Specifically built to implement the palatable units framework.

## References

1.  Spiegelhalter, D., & Blastland, M. (2013). *The Norm Chronicles: Stories and numbers about danger*. Profile Books.
2.  Spiegelhalter, D. (2019). *The Art of Statistics: Learning from Data*. Pelican.
