---
title: "Regional Variation in Life Expectancy"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Regional Variation in Life Expectancy}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  class.source = "fold-hide"
)
```

```{r setup}
library(micromort)
library(dplyr)
library(ggplot2)
library(DT)
```

This vignette explores **regional variation in life expectancy** across Western Europe, based on research by Bonnet et al. (2026) published in *Nature Communications*.

## Understanding the Data Structure

**Each row represents aggregated population statistics for one region-year-sex combination**, NOT individual survey responses.

| region_code | year | sex | life_expectancy | What this means |
|-------------|------|-----|-----------------|-----------------|
| FR10 | 2019 | Male | 82.5 | Average LE for **all males** in Île-de-France in 2019 |
| FR10 | 2019 | Female | 87.1 | Average LE for **all females** in Île-de-France in 2019 |
| FR10 | 2019 | Total | 84.8 | Average LE for **entire population** of Île-de-France in 2019 |

The underlying Eurostat data represents **~400 million people** across Western Europe. Life expectancy is calculated from official death registrations and census population counts—not a sample survey.

**Row count formula:** `regions × years × 3 sex categories`

- Sample data: 11 regions × 28 years × 3 = **924 rows**
- Full dataset: 450 regions × 28 years × 3 = **37,800 rows**

## Key Finding: A Two-Tiered Europe

Since the mid-2000s, Western Europe has fragmented into:
- **Vanguard regions**: Continued progress (~2.5 months/year gain for men)
- **Laggard regions**: Stalled improvement (<0.5 months/year gain)

This divergence reversed decades of convergence observed in the 1990s.

```{r}
# Compare life expectancy by classification (2019)
regional_life_expectancy(year = 2019, sex = "Total") |>
  group_by(classification) |>
  summarise(
    n_regions = n(),
    mean_le = round(mean(life_expectancy), 1),
    min_le = round(min(life_expectancy), 1),
    max_le = round(max(life_expectancy), 1),
    mean_microlives_diff = round(mean(microlives_vs_eu_avg), 1)
  ) |>
  DT::datatable(
    caption = "Life expectancy by region classification (2019)",
    options = list(dom = "t", pageLength = 5),
    rownames = FALSE
  )
```

## The Microlives Gap

The ~7 year gap between vanguard and laggard regions translates to a substantial lifetime difference in microlives:

```{r}
# Calculate microlives gap
gap_data <- regional_life_expectancy(year = 2019, sex = "Total") |>
  group_by(classification) |>
  summarise(mean_le = mean(life_expectancy)) |>
  filter(classification %in% c("vanguard", "laggard"))

le_gap <- diff(gap_data$mean_le)

# Convert to microlives
# 1 year = 365 days × 48 microlives/day = 17,520 microlives
lifetime_microlives_gap <- abs(le_gap) * 17520

cat("Life expectancy gap:", round(abs(le_gap), 1), "years\n")
cat("Lifetime microlives difference:", format(round(lifetime_microlives_gap), big.mark = ","), "\n")
cat("Daily microlives difference:", round(abs(le_gap) * 1.2, 1), "per day\n")
```

**Interpretation:** Living in a vanguard region vs a laggard region corresponds to ~`r round(abs(le_gap) * 1.2, 1)` microlives per day—roughly equivalent to the benefit of 30 minutes of daily exercise.

## Regional Data Explorer

```{r}
# Full data table
regional_life_expectancy(year = 2019, sex = "Total") |>
  select(region_name, country_code, life_expectancy, microlives_vs_eu_avg, classification) |>
  arrange(desc(life_expectancy)) |>
  DT::datatable(
    caption = "Regional life expectancy (2019, click to sort/filter)",
    filter = "top",
    options = list(pageLength = 10, scrollX = TRUE),
    rownames = FALSE
  )
```

## Trends Over Time

The divergence became pronounced after 2005:

```{r, fig.width=8, fig.height=5}
# Plot trends by classification
regional_life_expectancy(sex = "Total") |>
  group_by(year, classification) |>
  summarise(mean_le = mean(life_expectancy), .groups = "drop") |>
  ggplot(aes(x = year, y = mean_le, color = classification)) +
  geom_line(linewidth = 1.2) +
  geom_vline(xintercept = 2005, linetype = "dashed", alpha = 0.5) +
  annotate("text", x = 2006, y = 74, label = "Divergence\nbegins", hjust = 0, size = 3) +
  scale_color_manual(
    values = c("vanguard" = "#2E7D32", "average" = "#1976D2", "laggard" = "#C62828"),
    labels = c("Vanguard", "Average", "Laggard")
  ) +
  labs(
    title = "Life Expectancy Trends by Region Classification",
    subtitle = "Western Europe, 1992-2019",
    x = "Year",
    y = "Life Expectancy at Birth (years)",
    color = "Classification",
    caption = "Source: Eurostat demo_r_mlifexp; Classification per Bonnet et al. (2026)"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

## Mortality Risk Multiplier

Use `regional_mortality_multiplier()` to adjust baseline micromort estimates by location:

```{r}
# Compare Paris region to EU average
regional_mortality_multiplier("FR10") |>
  DT::datatable(
    caption = "Île-de-France (Paris) mortality multiplier",
    options = list(dom = "t"),
    rownames = FALSE
  )
```

**Application:** If the baseline risk for an activity is 10 micromorts, the location-adjusted risk in Paris would be approximately `10 × 0.93 = 9.3 micromorts` (7% lower due to favorable regional factors).

## Ecological Fallacy Warning

**IMPORTANT:** These regional statistics reflect population averages, not individual-level causation.

High life expectancy in "vanguard" regions results from multiple interacting factors:

| Factor | Mechanism |
|--------|-----------|
| Healthcare access | Better hospitals, preventive care |
| Socioeconomic composition | Higher income, education levels |
| Selection effects | Healthy/wealthy people move to desirable regions |
| Historical factors | Long-term infrastructure investments |
| Cultural factors | Diet, social cohesion, lifestyle norms |

**Moving to Switzerland will NOT automatically extend your life.** The regional advantage reflects the aggregate characteristics of people who already live there.

## Data Source

The regional classification methodology follows Bonnet et al. (2026):

> Bonnet F, et al. "Potential and challenges for sustainable progress in human longevity."
> *Nature Communications* 17, 996 (2026). [doi:10.1038/s41467-026-68828-z](https://doi.org/10.1038/s41467-026-68828-z)

Raw data from Eurostat `demo_r_mlifexp` dataset. Interactive exploration available at the [ReLoG_Europe tool](https://histdemo.shinyapps.io/ReLoG_Europe/).

## Functions Reference

| Function | Purpose |
|----------|---------|
| `regional_life_expectancy()` | Full dataset with filters |
| `vanguard_regions()` | Top-performing regions only |
| `laggard_regions()` | Stagnating regions only |
| `regional_mortality_multiplier()` | Location-based risk adjustment |
