---
title: "Regional Variation in Life Expectancy"
format: html
vignette: >
  %\VignetteIndexEntry{Regional Variation in Life Expectancy}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = FALSE
)

library(targets)
library(DT)

# Safe tar_read with graceful fallback
safe_tar_read <- function(name) {
  store <- Sys.getenv("TARGETS_STORE", unset = NA)
  if (is.na(store)) {
    store <- tryCatch(
      file.path(rprojroot::find_root(rprojroot::is_r_package), "_targets"),
      error = function(e) "_targets"
    )
  }
  tryCatch(
    targets::tar_read_raw(name, store = store),
    error = function(e) {
      message("Target '", name, "' not found. Run tar_make() first.")
      NULL
    }
  )
}
```

This vignette explores **regional variation in life expectancy** across Western Europe, based on research by Bonnet et al. (2026) published in *Nature Communications*.

## Understanding the Data Structure

**Each row represents aggregated population statistics for one region-year-sex combination**, NOT individual survey responses.

| region_code | year | sex | life_expectancy | What this means |
|-------------|------|-----|-----------------|-----------------|
| FR10 | 2019 | Male | 82.5 | Average LE for **all males** in Île-de-France in 2019 |
| FR10 | 2019 | Female | 87.1 | Average LE for **all females** in Île-de-France in 2019 |
| FR10 | 2019 | Total | 84.8 | Average LE for **entire population** of Île-de-France in 2019 |

The underlying Eurostat data represents **~400 million people** across Western Europe. Life expectancy is calculated from official death registrations and census population counts—not a sample survey.

**Row count formula:** `regions × years × 3 sex categories`

- Sample data: 11 regions × 28 years × 3 = **924 rows**
- Full dataset: 450 regions × 28 years × 3 = **37,800 rows**

## Key Finding: A Two-Tiered Europe

Since the mid-2000s, Western Europe has fragmented into:
- **Vanguard regions**: Continued progress (~2.5 months/year gain for men)
- **Laggard regions**: Stalled improvement (<0.5 months/year gain)

This divergence reversed decades of convergence observed in the 1990s.

```{r}
#| echo: false
safe_tar_read("vig_regional_classification_summary") |>
  DT::datatable(
    caption = "Life expectancy by region classification (2019)",
    options = list(dom = "t", pageLength = 5),
    rownames = FALSE
  )
```

## The Microlives Gap

The ~7 year gap between vanguard and laggard regions translates to a substantial lifetime difference in microlives:

```{r}
#| echo: false
gap_data <- safe_tar_read("vig_regional_le_gap")
if (!is.null(gap_data)) {
  cat("Life expectancy gap:", gap_data$le_gap_years, "years\n")
  cat("Lifetime microlives difference:", gap_data$lifetime_microlives, "\n")
  cat("Daily microlives difference:", gap_data$daily_microlives, "per day\n")
}
```

**Interpretation:** Living in a vanguard region vs a laggard region corresponds to ~`r if(!is.null(gap_data)) gap_data$daily_microlives else "8.4"` microlives per day—roughly equivalent to the benefit of 30 minutes of daily exercise.

## Regional Data Explorer

**Data period:** 2019 (pre-COVID baseline year, last year before pandemic distortions)

**Column definitions:**

| Column | Definition | Units |
|--------|------------|-------|
| `region_name` | NUTS2 administrative region | — |
| `country_code` | ISO 2-letter country code | — |
| `life_expectancy` | Period life expectancy at birth | Years |
| `microlives_vs_eu_avg` | Daily microlives gained/lost vs EU average | Microlives/day |
| `classification` | Vanguard (top 20% + growing), Laggard (bottom 20% or stagnant), Average | — |

**Key findings:**

- **Vanguard-laggard gap:** ~7 years LE difference = **~8.4 microlives/day** (equivalent to 30 min daily exercise)
- **Gap trend:** Widened from ~5 years (1992) to ~7 years (2019) as laggard regions stagnated post-2005
- **Top region:** Comunidad de Madrid (ES) at 86.1 years
- **Bottom region:** Mayotte (FR overseas) at 74.9 years
- **Microlives interpretation:** +1.0 microlives/day ≈ +30 min life expectancy/day ≈ +7.6 days/year

```{r}
#| echo: false
safe_tar_read("vig_regional_explorer_data") |>
  DT::datatable(
    caption = htmltools::HTML(
      "<b>Regional Life Expectancy at Birth (2019)</b><br>
      <small>Source: Eurostat demo_r_mlifexp. 229 NUTS2 regions across 16 Western European countries.<br>
      Period: 2019 (pre-COVID baseline). Sex: Total (both sexes combined).<br>
      Microlives: Daily life expectancy difference vs EU average (1 microlife = 30 min LE change per day).<br>
      Classification: Based on 2019 LE percentile and 5-year trend (Bonnet et al. 2026 methodology).</small>"
    ),
    filter = "top",
    options = list(pageLength = 15, scrollX = TRUE),
    rownames = FALSE
  )
```

## Trends Over Time

The divergence became pronounced after 2005:

```{r, fig.width=8, fig.height=5}
#| echo: false
print(safe_tar_read("vig_regional_trends_plot"))
```

## Mortality Risk Multiplier

Use `regional_mortality_multiplier()` to adjust baseline micromort estimates by location:

```{r}
#| echo: false
safe_tar_read("vig_regional_paris_multiplier") |>
  DT::datatable(
    caption = "Île-de-France (Paris) mortality multiplier",
    options = list(dom = "t"),
    rownames = FALSE
  )
```

**Application:** If the baseline risk for an activity is 10 micromorts, the location-adjusted risk in Paris would be approximately `10 × 0.93 = 9.3 micromorts` (7% lower due to favorable regional factors).

## Ecological Fallacy Warning

**IMPORTANT:** These regional statistics reflect population averages, not individual-level causation.

High life expectancy in "vanguard" regions results from multiple interacting factors:

| Factor | Mechanism |
|--------|-----------|
| Healthcare access | Better hospitals, preventive care |
| Socioeconomic composition | Higher income, education levels |
| Selection effects | Healthy/wealthy people move to desirable regions |
| Historical factors | Long-term infrastructure investments |
| Cultural factors | Diet, social cohesion, lifestyle norms |

**Moving to Switzerland will NOT automatically extend your life.** The regional advantage reflects the aggregate characteristics of people who already live there.

## Data Source

The regional classification methodology follows Bonnet et al. (2026):

> Bonnet F, et al. "Potential and challenges for sustainable progress in human longevity."
> *Nature Communications* 17, 996 (2026). [doi:10.1038/s41467-026-68828-z](https://doi.org/10.1038/s41467-026-68828-z)

Raw data from Eurostat `demo_r_mlifexp` dataset. Interactive exploration available at the [ReLoG_Europe tool](https://histdemo.shinyapps.io/ReLoG_Europe/).

## Functions Reference

| Function | Purpose |
|----------|---------|
| `regional_life_expectancy()` | Full dataset with filters |
| `vanguard_regions()` | Top-performing regions only |
| `laggard_regions()` | Stagnating regions only |
| `regional_mortality_multiplier()` | Location-based risk adjustment |
