---
title: "REST API"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{REST API}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{css, echo=FALSE}
.main-container { max-width: 95vw !important; }
body .container-fluid { max-width: 95vw; }
```

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  class.source = "fold-hide"
)
```

```{r setup, include=FALSE}
library(micromort)
library(targets)
library(DT)

safe_tar_read <- function(name) {
  store <- Sys.getenv("TARGETS_STORE", unset = NA)
  if (is.na(store)) {
    store <- tryCatch(
      file.path(rprojroot::find_root(rprojroot::is_r_package), "_targets"),
      error = function(e) "_targets"
    )
  }
  tryCatch(
    targets::tar_read_raw(name, store = store),
    error = function(e) {
      message("Target '", name, "' not found. Run tar_make() first.")
      NULL
    }
  )
}
```

The micromort package includes a full REST API with 30 endpoints for accessing
risk data programmatically. The API is built with [plumber](https://www.rplumber.io/)
and returns JSON with a standard response envelope.

## Getting Started

Launch the API from R:

```r
library(micromort)
launch_api()
#> ── Micromort Data API ──
#> ℹ Starting server at http://127.0.0.1:8080
#> ℹ Swagger docs: http://127.0.0.1:8080/__docs__/
```

The interactive Swagger UI at `/__docs__/` lets you explore all endpoints,
view parameter descriptions, and try requests directly in the browser.

## Response Envelope

Every endpoint returns a standard JSON envelope:

```json
{
  "data": [ ... ],
  "meta": {
    "source": "micromort v0.4.0",
    "endpoint": "/v1/risks/acute",
    "n_rows": 10,
    "timestamp": "2026-03-09T12:00:00Z",
    "params": { "category": "Medical" }
  }
}
```

The `data` field contains the result (array or object). The `meta` field includes
package version, endpoint path, row count, ISO 8601 timestamp, and the query
parameters used.

## Core Risk Data

### Acute risks

Retrieve enriched acute risk data with optional filtering by category,
minimum micromort threshold, and result limit.

```bash
curl "http://localhost:8080/v1/risks/acute?category=Medical&limit=10"
```

```{r acute-sample}
acute <- safe_tar_read("vig_api_acute_sample")
if (!is.null(acute)) {
  DT::datatable(
    acute,
    caption = "GET /v1/risks/acute?category=Medical — Medical activities ranked by risk",
    options = list(pageLength = 10, dom = "tip", scrollX = TRUE),
    rownames = FALSE
  ) |>
    DT::formatRound(c("micromorts", "microlives"), 2)
}
```

### Chronic risks

Query chronic lifestyle factors that gain or lose microlives per day.

```bash
curl "http://localhost:8080/v1/risks/chronic?direction=gain"
```

```{r chronic-gains}
chronic <- safe_tar_read("vig_api_chronic_gains")
if (!is.null(chronic)) {
  DT::datatable(
    chronic,
    caption = "GET /v1/risks/chronic?direction=gain — Factors that extend life",
    options = list(pageLength = 15, dom = "tip", scrollX = TRUE),
    rownames = FALSE
  ) |>
    DT::formatRound("microlives_per_day", 1)
}
```

### Cancer risks

Cancer mortality by type, sex, and age group with family history multipliers.

```bash
curl "http://localhost:8080/v1/risks/cancer?age_group=All%20ages"
```

```{r cancer-top3}
cancer <- safe_tar_read("vig_api_cancer_top3")
if (!is.null(cancer)) {
  DT::datatable(
    cancer,
    caption = "GET /v1/risks/cancer — Top 3 cancers per sex (All ages)",
    options = list(pageLength = 10, dom = "tip", scrollX = TRUE),
    rownames = FALSE
  ) |>
    DT::formatRound(c("deaths_per_100k", "micromorts_per_year"), 1)
}
```

## Risk Analysis

### Risk equivalence

Find activities with equivalent risk to a reference activity. The response
includes the ratio (how many of the reference activity equals one of each
comparison activity).

```bash
curl "http://localhost:8080/v1/analysis/equivalence?reference=Chest+X-ray+(radiation+per+scan)"
```

```{r equivalence-sample}
equiv <- safe_tar_read("vig_api_equivalence_sample")
if (!is.null(equiv)) {
  DT::datatable(
    equiv,
    caption = "GET /v1/analysis/equivalence — Activities expressed in chest X-ray equivalents",
    options = list(pageLength = 15, dom = "tip", scrollX = TRUE),
    rownames = FALSE
  ) |>
    DT::formatRound("ratio", 1)
}
```

### Exchange matrix

Build a cross-comparison matrix for multiple activities. This is a POST
endpoint that accepts a JSON body with an optional `activities` array:

```bash
curl -X POST "http://localhost:8080/v1/analysis/exchange-matrix" \
  -H "Content-Type: application/json" \
  -d '{"activities": ["Skiing", "Scuba diving, trained", "Skydiving (US)"]}'
```

## Unit Conversion

Convert between probability, micromorts, microlives, and loss of life expectancy.

```bash
# Probability to micromorts
curl "http://localhost:8080/v1/convert/to-micromort?prob=0.000001"

# Micromorts to probability
curl "http://localhost:8080/v1/convert/to-probability?micromorts=1"

# Loss of life expectancy (minutes)
curl "http://localhost:8080/v1/convert/lle?prob=0.00001&life_expectancy=40"

# Monetary value of a micromort (default VSL = $10M)
curl "http://localhost:8080/v1/convert/value?vsl=10000000"
```

```{r conversion-table}
conversions <- safe_tar_read("vig_api_conversion_table")
if (!is.null(conversions)) {
  DT::datatable(
    conversions,
    caption = "Unit conversions across the probability-to-microlife spectrum",
    options = list(pageLength = 10, dom = "tip", scrollX = TRUE),
    rownames = FALSE
  ) |>
    DT::formatSignif(c("probability", "micromorts", "lle_minutes", "microlife"), 3)
}
```

## Age-Based Hazard Rates

Calculate daily micromort exposure from background mortality at any age,
using Gompertz-Makeham mortality models.

```bash
curl "http://localhost:8080/v1/convert/hazard-rate?age=50&sex=female"
```

```{r hazard-ages}
hazard <- safe_tar_read("vig_api_hazard_ages")
if (!is.null(hazard)) {
  DT::datatable(
    hazard,
    caption = "GET /v1/convert/hazard-rate — Daily background mortality by age and sex",
    options = list(pageLength = 12, dom = "tip", scrollX = TRUE),
    rownames = FALSE
  ) |>
    DT::formatRound("micromorts", 1) |>
    DT::formatSignif("daily_prob", 3)
}
```

## Full Endpoint Reference

All 30 API endpoints with their HTTP method, path, description, and parameters:

```{r endpoint-summary}
endpoints <- safe_tar_read("vig_api_endpoint_summary")
if (!is.null(endpoints)) {
  DT::datatable(
    endpoints,
    caption = "Complete API endpoint reference",
    options = list(
      pageLength = 30,
      dom = "ftip",
      columnDefs = list(list(width = "250px", targets = 1))
    ),
    rownames = FALSE,
    filter = "top"
  )
}
```

## Using from R

You can call the API from R using [httr2](https://httr2.r-lib.org/):

```{r httr2-example, eval=FALSE}
library(httr2)

base_url <- "http://localhost:8080"

# GET request with query parameters
resp <- request(base_url) |>
  req_url_path("/v1/risks/acute") |>
  req_url_query(category = "Sport", limit = 5) |>
  req_perform() |>
  resp_body_json()

# The data lives in resp$data
tibble::as_tibble(do.call(rbind, lapply(resp$data, as.data.frame)))

# POST request with JSON body
resp <- request(base_url) |>
  req_url_path("/v1/analysis/exchange-matrix") |>
  req_body_json(list(
    activities = c("Skiing", "Scuba diving, trained")
  )) |>
  req_perform() |>
  resp_body_json()
```
