---
title: "Risk Equivalence Dashboard"
format: html
vignette: >
  %\VignetteIndexEntry{Risk Equivalence Dashboard}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{css, echo=FALSE}
/* Use 95% of viewport width */
.main-container { max-width: 95vw !important; }
body .container-fluid { max-width: 95vw; }
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = FALSE
)

if (requireNamespace("micromort", quietly = TRUE)) {
  library(micromort)
} else {
  pkgload::load_all(quiet = TRUE)
}

library(targets)
library(DT)

# Safe tar_read with graceful fallback
safe_tar_read <- function(name) {
  store <- Sys.getenv("TARGETS_STORE", unset = NA)
  if (is.na(store)) {
    store <- tryCatch(
      file.path(rprojroot::find_root(rprojroot::is_r_package), "_targets"),
      error = function(e) "_targets"
    )
  }
  tryCatch(
    targets::tar_read_raw(name, store = store),
    error = function(e) {
      message("Target '", name, "' not found. Run tar_make() first.")
      NULL
    }
  )
}
```

Micromorts provide a **common currency** for comparing risks that otherwise seem incommensurable. How dangerous is a CT scan compared to a skydive? How many chest X-rays equal a long-haul flight?

This dashboard explores these questions using **atomic risk decomposition** --- breaking composite activities into their individual risk components so you can see exactly what you're exposed to, and what you can mitigate.

## Everyday Risk Budget {.tabset}

### Table

How risky are the mundane things we do every day?

```{r everyday-table}
everyday <- safe_tar_read("vig_equiv_everyday")
if (!is.null(everyday)) {
  DT::datatable(
    everyday,
    caption = "Everyday activities ranked by micromort risk, with chest X-ray equivalents",
    options = list(pageLength = 15, dom = "tip", scrollX = TRUE),
    rownames = FALSE
  ) |>
    DT::formatRound(c("micromorts", "microlives", "micromorts_per_day"), 2)
}
```

### Chart

Everyday activities expressed in chest X-ray equivalents:

```{r everyday-chart, fig.width=8, fig.height=5}
#| fig-cap: "Everyday activities expressed in chest X-ray equivalents, showing that a long-haul flight equals ~50 chest X-rays while a banana is negligible."
#| fig-alt: "Horizontal bar chart comparing everyday activities to chest X-ray dose equivalents. Activities ordered from least to most radiation-equivalent, with blue bars."
everyday <- safe_tar_read("vig_equiv_everyday")
if (!is.null(everyday) && requireNamespace("plotly", quietly = TRUE)) {
  plotly::plot_ly(
    everyday,
    y = ~stats::reorder(activity, xray_equivalents),
    x = ~xray_equivalents,
    type = "bar",
    orientation = "h",
    marker = list(color = "#1976D2")
  ) |>
    plotly::layout(
      title = "Everyday Activities in Chest X-ray Equivalents",
      xaxis = list(title = "Chest X-ray equivalents"),
      yaxis = list(title = ""),
      margin = list(l = 200),
      paper_bgcolor = "white",
      plot_bgcolor = "white",
      font = list(color = "#1a1a1a")
    ) |>
    plotly::config(scrollZoom = TRUE)
}
```


## Flight Risk Decomposition {.tabset}

Flying is a composite risk: crash + deep vein thrombosis (DVT) + cosmic radiation. The **atomic decomposition** reveals which components dominate at each duration, and which you can mitigate.

### By Duration

```{r flight-duration, fig.width=9, fig.height=5}
#| fig-cap: "Stacked bar chart decomposing flight risk into crash, DVT, and cosmic radiation components across 2h, 5h, 8h, and 12h flights."
#| fig-alt: "Stacked bar chart showing flight micromorts by duration. Crash risk is constant (~1 mm), DVT appears above 4 hours and grows nonlinearly, cosmic radiation increases linearly with duration."
flight_data <- safe_tar_read("vig_equiv_flight_duration")
if (!is.null(flight_data) && requireNamespace("plotly", quietly = TRUE)) {
  plotly::plot_ly(
    flight_data,
    x = ~activity,
    y = ~micromorts,
    color = ~component_label,
    type = "bar"
  ) |>
    plotly::layout(
      barmode = "stack",
      title = "Flight Risk by Duration and Component (Healthy Profile)",
      xaxis = list(title = ""),
      yaxis = list(title = "Micromorts"),
      legend = list(orientation = "h", y = -0.2, font = list(color = "#1a1a1a"),
                    bgcolor = "rgba(255,255,255,0.9)"),
      paper_bgcolor = "white",
      plot_bgcolor = "white",
      font = list(color = "#1a1a1a")
    ) |>
    plotly::config(scrollZoom = TRUE)
}
```

Key observations:

- **Crash risk** is roughly constant per flight (~1 mm) regardless of duration --- dominated by takeoff and landing phases (~80% of fatal accidents per Boeing Statistical Summary) --- and is NOT hedgeable
- **DVT risk** is zero below 4 hours, then grows nonlinearly --- and IS hedgeable (compression socks reduce risk ~65%)
- **Cosmic radiation** is linear (~0.05 mm/hour) and NOT hedgeable
- For an 8-hour flight, DVT is the dominant hedgeable component

### By Health Status

How does DVT risk status change the total?

```{r flight-health}
components <- safe_tar_read("vig_equiv_flight_components")
if (!is.null(components)) {
  DT::datatable(
    components,
    caption = "Flying (8h long-haul): Healthy vs DVT risk factors",
    options = list(pageLength = 10, dom = "tip", scrollX = TRUE),
    rownames = FALSE
  ) |>
    DT::formatRound("micromorts", 1)
}
```

### Components

```{r flight-components}
# Show all flight components across durations
flight_data <- safe_tar_read("vig_equiv_flight_duration")
if (!is.null(flight_data)) {
  DT::datatable(
    flight_data,
    caption = "All flight risk components by duration (healthy profile)",
    options = list(pageLength = 20, dom = "tip", scrollX = TRUE),
    rownames = FALSE
  ) |>
    DT::formatRound("micromorts", 2)
}
```


## Landmark Equivalences {.tabset}

### Surprise Table

```{r landmarks}
landmarks <- safe_tar_read("vig_equiv_landmarks")
if (!is.null(landmarks)) {
  DT::datatable(
    landmarks |>
      dplyr::select(activity, micromorts, category, xray_equivalents),
    caption = "Landmark activities from coffee to Everest, in chest X-ray equivalents",
    options = list(pageLength = 20, dom = "tip", scrollX = TRUE),
    rownames = FALSE
  ) |>
    DT::formatRound("micromorts", 2)
}
```

### Interactive Explorer

Full risk equivalence table with every activity expressed relative to a chest X-ray:

```{r explorer}
explorer <- safe_tar_read("vig_equiv_explorer")
if (!is.null(explorer)) {
  DT::datatable(
    explorer,
    caption = "All activities as multiples of one chest X-ray (0.1 micromorts)",
    options = list(pageLength = 15, dom = "ftip", scrollX = TRUE),
    filter = "top",
    rownames = FALSE
  ) |>
    DT::formatRound(c("micromorts", "ratio"), 2)
}
```


## Medical Radiation {.tabset}

### Comparison Table

Medical imaging procedures vary enormously in radiation dose:

```{r medical-radiation}
med <- safe_tar_read("vig_equiv_medical_focus")
if (!is.null(med)) {
  DT::datatable(
    med |> dplyr::select(activity, micromorts, microlives, period),
    caption = "Medical radiation procedures ranked by micromort risk",
    options = list(pageLength = 10, dom = "tip", scrollX = TRUE),
    rownames = FALSE
  ) |>
    DT::formatRound(c("micromorts", "microlives"), 2)
}
```

### Exchange Chart

How many chest X-rays equal one CT scan?

```{r medical-exchange, fig.width=8, fig.height=4}
#| fig-cap: "Medical procedures ranked by chest X-ray equivalents, showing that a CT abdomen equals ~200 chest X-rays."
#| fig-alt: "Horizontal bar chart of medical radiation procedures in chest X-ray equivalents. Red bars show procedures from dental X-ray (lowest) to CT abdomen (highest, ~200 equivalents)."
med <- safe_tar_read("vig_equiv_medical_focus")
if (!is.null(med) && requireNamespace("plotly", quietly = TRUE)) {
  med <- med |>
    dplyr::mutate(xray_equiv = round(micromorts / 0.1, 0))

  plotly::plot_ly(
    med,
    y = ~stats::reorder(activity, xray_equiv),
    x = ~xray_equiv,
    type = "bar",
    orientation = "h",
    marker = list(color = "#C62828")
  ) |>
    plotly::layout(
      title = "Medical Procedures in Chest X-ray Equivalents",
      xaxis = list(title = "Number of chest X-rays"),
      yaxis = list(title = ""),
      margin = list(l = 200),
      paper_bgcolor = "white",
      plot_bgcolor = "white",
      font = list(color = "#1a1a1a")
    ) |>
    plotly::config(scrollZoom = TRUE)
}
```


## Hedgeability Analysis {.tabset}

### By Activity

Which activities have hedgeable risk components?

```{r hedgeable}
hedgeable <- safe_tar_read("vig_equiv_hedgeable_summary")
if (!is.null(hedgeable)) {
  DT::datatable(
    hedgeable,
    caption = "Activities with hedgeable risk components (sorted by hedgeable %)",
    options = list(pageLength = 10, dom = "tip", scrollX = TRUE),
    rownames = FALSE
  ) |>
    DT::formatRound("hedgeable_pct", 1)
}
```

### Stacked Components

Flight risk decomposition showing hedgeable vs non-hedgeable portions:

```{r hedgeable-plot, fig.width=9, fig.height=5}
#| fig-cap: "Flight risk decomposition by hedgeability: DVT risk (green, hedgeable via compression socks) vs crash and radiation (red, not hedgeable)."
#| fig-alt: "Stacked bar chart of flight micromorts by duration, coloured green for hedgeable components (DVT) and red for non-hedgeable (crash, cosmic radiation). DVT dominates hedgeable risk on longer flights."
flight_data <- safe_tar_read("vig_equiv_flight_duration")
if (!is.null(flight_data) && requireNamespace("plotly", quietly = TRUE)) {
  flight_data <- flight_data |>
    dplyr::mutate(
      hedge_status = ifelse(hedgeable, "Hedgeable", "Not hedgeable")
    )

  plotly::plot_ly(
    flight_data,
    x = ~activity,
    y = ~micromorts,
    color = ~hedge_status,
    colors = c("Hedgeable" = "#2E7D32", "Not hedgeable" = "#C62828"),
    type = "bar"
  ) |>
    plotly::layout(
      barmode = "stack",
      title = "Hedgeable vs Non-hedgeable Risk by Flight Duration",
      xaxis = list(title = ""),
      yaxis = list(title = "Micromorts"),
      legend = list(orientation = "h", y = -0.2, font = list(color = "#1a1a1a"),
                    bgcolor = "rgba(255,255,255,0.9)"),
      paper_bgcolor = "white",
      plot_bgcolor = "white",
      font = list(color = "#1a1a1a")
    ) |>
    plotly::config(scrollZoom = TRUE)
}
```


## Radiation Exposure Profiles {.tabset}

How does occupational radiation exposure compare across careers, and how do patient X-ray doses stack up? This section uses annual dose data from UNSCEAR and the LNT model (0.05 micromorts per mSv) to answer these questions.

### Occupational Comparison

Annual and cumulative radiation exposure across 11 profiles --- occupational, passenger, and environmental:

```{r radiation-profiles}
rad_profiles <- safe_tar_read("vig_radiation_profiles")
if (!is.null(rad_profiles)) {
  DT::datatable(
    rad_profiles,
    caption = "Radiation exposure profiles: annual dose and cumulative micromorts at career milestones",
    options = list(pageLength = 15, dom = "tip", scrollX = TRUE),
    rownames = FALSE
  ) |>
    DT::formatRound(c("annual_msv", "annual_micromorts",
                       "mm_10y", "mm_20y", "mm_40y",
                       "xray_equivalents_per_year"), 3)
}
```

Key insight: A 40-year airline pilot career accumulates ~6 micromorts of radiation --- equivalent to just 60 chest X-rays.

### Patient vs Occupational

How many patient X-rays equal a career of occupational exposure?

```{r radiation-patient-vs-occ}
prc <- safe_tar_read("vig_radiation_patient_vs_occ")
if (!is.null(prc)) {
  DT::datatable(
    prc,
    caption = "Patient X-ray exposure vs occupational career radiation (ratio > 1 means patient exceeds worker)",
    options = list(pageLength = 15, dom = "tip", scrollX = TRUE),
    filter = "top",
    rownames = FALSE
  ) |>
    DT::formatRound(c("patient_micromorts", "occupational_micromorts", "ratio"), 3)
}
```

Key insight: 100 lifetime chest X-rays (10 micromorts) exceeds a 40-year X-ray technician career (2 micromorts) by 5x.

### Timeline

Cumulative radiation exposure over a 40-year career:

```{r radiation-timeline, fig.width=9, fig.height=6}
#| fig-cap: "Cumulative radiation micromorts over a 40-year career for different exposure profiles, showing that 100 lifetime chest X-rays exceeds occupational exposure."
#| fig-alt: "Line chart of cumulative micromorts vs years of exposure. Multiple lines for different radiation scenarios (dental, X-ray tech, CT patient, annual cosmic). Lines diverge over time showing compounding differences."
timeline <- safe_tar_read("vig_radiation_timeline_data")
if (!is.null(timeline) && requireNamespace("plotly", quietly = TRUE)) {
  plotly::plot_ly(
    timeline,
    x = ~year,
    y = ~cumulative_micromorts,
    color = ~activity,
    type = "scatter",
    mode = "lines"
  ) |>
    plotly::layout(
      title = "Cumulative Radiation Micromorts Over Career",
      xaxis = list(title = "Years of Exposure"),
      yaxis = list(title = "Cumulative Micromorts"),
      legend = list(orientation = "v", x = 1.02, y = 1, font = list(color = "#1a1a1a"),
                    bgcolor = "rgba(255,255,255,0.9)"),
      paper_bgcolor = "white",
      plot_bgcolor = "white",
      font = list(color = "#1a1a1a")
    ) |>
    plotly::config(scrollZoom = TRUE)
}
```

### Regulatory Context

How do actual doses compare to ICRP regulatory limits?

```{r radiation-regulatory}
regulatory <- safe_tar_read("vig_radiation_regulatory")
if (!is.null(regulatory)) {
  DT::datatable(
    regulatory,
    caption = "Actual annual doses vs ICRP limits (occupational: 20 mSv/yr, public: 1 mSv/yr)",
    options = list(pageLength = 15, dom = "tip", scrollX = TRUE),
    rownames = FALSE
  ) |>
    DT::formatRound(c("annual_msv", "pct_of_limit"), 1)
}
```

Key insight: Actual doses are typically 5-20x below regulatory limits.


## Cross-Activity Matrix

Exchange rates between 10 diverse activities. Read as: "one row-activity equals X column-activities."

```{r matrix}
mat <- safe_tar_read("vig_equiv_matrix")
if (!is.null(mat)) {
  DT::datatable(
    mat,
    caption = "Risk exchange matrix: cell(i,j) = how many of column j equal one of row i",
    options = list(
      pageLength = 10,
      dom = "t",
      scrollX = TRUE
    ),
    rownames = FALSE
  ) |>
    DT::formatRound(names(mat)[-1], 1)
}
```


## Methodology & Caveats

**Atomic vs composite risks.** The `atomic_risks()` function returns ONE row per risk component per activity. Activities not yet decomposed use `component = "all_causes"` (an honest placeholder indicating the breakdown is unknown).

**Conditional risks.** Some components depend on health profile (e.g., DVT risk varies by whether you have risk factors). The default profile assumes "healthy" values; use `common_risks(profile = list(health_profile = "dvt_risk_factors"))` for alternatives. Geographic conditioning can change equivalences dramatically: a snake bite is 0.5 mm in the US but 18.5 mm in rural Africa (37x). See the [Data Reliability](data_reliability.html) vignette for details.

**Duration bucketing.** Rather than encoding rate functions, flight risks are pre-computed at standard duration buckets (2h, 5h, 8h, 12h). Every number is directly citable --- no hidden formulas.

**DVT literature.** DVT risk below 4 hours is negligible. Above 4 hours, risk rises nonlinearly. Compression socks + hydration + movement reduce DVT risk by approximately 60--70%. Sources: Lancet Haematology, Cochrane Reviews.

**Medical radiation.** The "(radiation)" label indicates that the radiation dose IS the risk. For invasive procedures (e.g., coronary angiogram), procedural risks (infection, bleeding) are separate components not yet decomposed.

**Confidence levels.** Each component carries a confidence rating: "high" (published meta-analyses), "medium" (single studies or expert consensus), "low" (extrapolated), "estimated" (order-of-magnitude).
