---
title: "What is a Micromort?"
subtitle: "A visual introduction to everyday risk — then test yourself"
format:
  closeread-html:
    code-fold: true
    code-summary: "Show code"
    remove-header-space: true
    css: closeread-standalone.css
vignette: >
  %\VignetteIndexEntry{What is a Micromort?}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", echo = FALSE,
  fig.width = 8, fig.height = 5, dpi = 150,
  warning = FALSE, message = FALSE
)
if (requireNamespace("micromort", quietly = TRUE)) {
  library(micromort)
} else {
  pkgload::load_all(quiet = TRUE)
}
library(ggplot2)
library(dplyr)

dark_theme <- theme_minimal(base_size = 15) +
  theme(
    plot.background = element_rect(fill = "#1a1a2e", colour = NA),
    panel.background = element_rect(fill = "#1a1a2e", colour = NA),
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.major.x = element_line(colour = "#333355"),
    text = element_text(colour = "#e0e0e0"),
    axis.text = element_text(colour = "#e0e0e0"),
    plot.caption = element_text(colour = "#888888", size = 11),
    legend.background = element_rect(fill = "#1a1a2e"),
    legend.text = element_text(colour = "#e0e0e0"),
    legend.title = element_text(colour = "#e0e0e0")
  )
```

```{css}
.cr-section .narrative-col p,
.cr-section .narrative-col li {
  font-size: 1.25rem;
  line-height: 1.8;
}
.cr-section .narrative-col h3 {
  font-size: 1.6rem;
}
```

:::{.cr-section}

### How dangerous is your morning?

Every morning involves risk. Brewing coffee. Crossing the road. Commuting to work. Each activity carries a measurable probability of death.[@cr-mundane]{focus-on="cr-mundane"}

A **micromort (mm)** is a one-in-a-million chance of death from a single activity.

| Activity | Micromorts (mm) | Ratio vs coffee |
|----------|:---------:|:---------:|
| Cup of coffee | 0.01 | 1x |
| Crossing a road | 0.02 | 2x |
| Office work (8 hrs) | 0.03 | 3x |
| Taking a bath | 0.07 | 7x |
| Commuting by car | 0.13 | 13x |
| Glass of wine | 0.50 | 50x |
| Commuting by bicycle | 0.50 | 50x |

Cycling to work is **50 times riskier than drinking a cup of coffee** — and roughly **4 times riskier than driving** (0.50 / 0.13 = 3.8x).

:::{#cr-mundane}
```{r mundane-plot}
cr <- common_risks()
mundane <- cr |>
  filter(activity %in% c(
    "Cup of coffee", "Crossing a road",
    "Commuting by car (30 min)", "Commuting by bicycle (30 min)",
    "Taking a bath", "Working in an office (8 hours)",
    "Drinking a glass of wine")) |>
  arrange(micromorts) |>
  mutate(activity = factor(activity, levels = activity))

ggplot(mundane, aes(x = micromorts, y = activity)) +
  geom_col(fill = "#2c7be5", alpha = 0.9) +
  geom_text(aes(label = sprintf("%.2f", micromorts)),
            hjust = -0.1, size = 4.5, colour = "#e0e0e0") +
  scale_x_continuous(expand = expansion(mult = c(0, 0.3))) +
  labs(x = "Micromorts (mm)", y = NULL,
       caption = "1 mm = one-in-a-million chance of death per activity") +
  dark_theme
```
:::

:::


:::{.cr-section}

### The full spectrum

Now zoom out.[@cr-spectrum]{focus-on="cr-spectrum"}

The micromort scale spans **six orders of magnitude** — from a dental X-ray (0.05 mm) to summiting Everest (37,932 mm).

**Bottom** (< 1 mm): daily activities — commuting, coffee, office work, dental X-rays. Risks you never notice.

**Middle** (1–100 mm): risks you weigh consciously — a night in hospital (75 mm), general anaesthesia (10 mm), a long-haul flight (4 mm), running a marathon (7 mm).

**Top** (> 100 mm): risks that make headlines — base jumping (430 mm per jump), Himalayan mountaineering (12,000 mm), **Everest at 37,932 mm — roughly a 4% chance of death per ascent**.

:::{#cr-spectrum}
```{r spectrum-plot, fig.height=10}
# Include items referenced in text + top risks
text_items <- c("Dental X-ray (radiation per scan)", "Night in hospital",
  "General anesthesia (emergency)", "Flying (8h long-haul)",
  "Running a marathon", "Base jumping (per jump)",
  "Himalayan mountaineering", "Mt. Everest ascent")

cr_show <- cr |>
  filter(micromorts > 0) |>
  arrange(desc(micromorts)) |>
  slice_head(n = 35) |>
  bind_rows(cr |> filter(activity %in% text_items)) |>
  distinct(activity, .keep_all = TRUE) |>
  arrange(desc(micromorts)) |>
  slice_head(n = 35) |>
  mutate(
    activity = factor(activity, levels = rev(activity)),
    highlight = ifelse(activity %in% text_items, "referenced", "other")
  )

ggplot(cr_show, aes(x = micromorts, y = activity)) +
  geom_point(aes(colour = category), size = 3) +
  geom_text(
    data = cr_show |> filter(highlight == "referenced"),
    aes(label = activity), hjust = -0.1, size = 3.5,
    colour = "#ff6b6b", fontface = "bold", nudge_x = 0.15) +
  scale_x_log10(labels = scales::label_comma()) +
  scale_colour_viridis_d(option = "turbo") +
  labs(x = "Micromorts, mm (log scale)", y = NULL, colour = "Category",
       caption = "Source: micromorts.rip, Wikipedia, CDC MMWR.\nRed labels = items referenced in text.") +
  dark_theme +
  theme(axis.text.y = element_text(size = 11))
```
:::

:::


:::{.cr-section}

### The slow burn

Not all risks kill in an instant.[@cr-chronic]{focus-on="cr-chronic"}

Some shorten your life by half an hour, every single day. David Spiegelhalter called these **microlives (ml)** — each one worth 30 minutes of life expectancy.

**Smoking 20 cigarettes a day costs 10 microlives.** That is 10 × 30 = 300 minutes = **5 hours of life, lost every 24 hours.** A heavy smoker lives as though the day were only 19 hours long.

The scale works in reverse too. **150 minutes of weekly exercise gains 3 ml/day.** Five servings of fruit and vegetables: **+4 ml/day** — yes, diet can outweigh exercise in Spiegelhalter's estimates (both from the BMJ 2012 paper). A Mediterranean diet: +2 ml/day.

The slow burn is not all loss. Some of it is gain.

:::{#cr-chronic}
```{r chronic-plot, fig.height=9}
ch <- chronic_risks()
ch <- ch |>
  arrange(microlives_per_day) |>
  mutate(
    factor = factor(factor, levels = factor),
    fill_colour = ifelse(direction == "gain", "#198754", "#dc3545")
  )

ggplot(ch, aes(x = microlives_per_day, y = factor)) +
  geom_col(aes(fill = fill_colour), alpha = 0.9) +
  scale_fill_identity() +
  geom_vline(xintercept = 0, colour = "#e0e0e0", linewidth = 0.5) +
  labs(x = "Microlives per day (ml/day)", y = NULL,
       caption = "1 ml = 30 min of life expectancy per day.\nRed = life lost. Green = life gained.") +
  dark_theme +
  theme(axis.text.y = element_text(size = 10))
```
:::

:::


:::{.cr-section}

### The bridge

How do you compare a skydive to a cigarette? One kills instantly (rarely). The other kills slowly (certainly, at scale).[@cr-bridge]{focus-on="cr-bridge"}

The answer is **Loss of Life Expectancy (LLE)** — the average time lost or gained per exposure.

One micromort costs roughly **21 minutes** of life expectancy. One microlife is **30 minutes**. So 1 mm ≈ 0.7 ml.

The UK Department for Transport values a micromort at **£1.60**. NICE values a microlife at **£1.70**. Two independent government agencies, nearly the same price for a unit of risk.

:::{#cr-bridge}

| Unit | Abbreviation | Definition | Example |
|------|:---:|-----------|---------|
| **Micromort** | mm | 1-in-a-million death probability | Skydiving (per jump) = 10 mm |
| **Microlife** | ml/day | 30 min of life expectancy per day | Smoking 2 cigs/day = −1 ml/day |
| **LLE** | min | Loss of Life Expectancy | 1 mm ≈ 21 min ≈ 0.7 ml |
| **£ value** | £ | NICE/DfT policy valuation | 1 mm ≈ £1.60 ≈ 1 ml/day |

: The three currencies of risk and the exchange rate between them. {.striped}

:::

:::


:::{.cr-section}

### Now test yourself

You now have three currencies for risk — micromorts (mm), microlives (ml/day), and loss of life expectancy (LLE).[@cr-quiz]{focus-on="cr-quiz"}

You know that coffee is 0.01 mm and Everest is 37,932 mm. You know that smoking costs 10 ml/day and exercise gains 3 ml/day. But how good is your intuition? Three quizzes, three formats:

**☠️ Micromort Quiz** — which activity is riskier? Pick A or B. Covers acute risks from skydiving to surgery.

**⏳ Microlife Quiz** — which daily habit has the bigger effect on your lifespan? Smoking vs diet vs exercise.

**📊 Rank Risks Quiz** — drag and drop 4 items into the correct order. Mixes acute and chronic risks on the same scale.

:::{#cr-quiz}

::: {style="background:#2b3e50;border-radius:12px;padding:28px;color:#fff;text-align:center;"}

### Ready?

[**☠️ Micromort Quiz**](micromort-quiz.html){.btn .btn-danger .btn-lg style="margin:8px;display:block;"}

Which event is more likely to kill you?

[**⏳ Microlife Quiz**](microlife-quiz.html){.btn .btn-success .btn-lg style="margin:8px;display:block;"}

Which daily habit has the bigger effect?

[**📊 Rank Risks Quiz**](risk-ranking-quiz.html){.btn .btn-primary .btn-lg style="margin:8px;display:block;"}

Rank 4 risks from biggest to smallest effect.

:::

:::

:::

<details>
<summary>Session Info</summary>
```{r session-info, eval=TRUE}
sessionInfo()
```
</details>
