---
title: "The Risk Landscape"
format:
  closeread-html:
    code-fold: true
    code-summary: "Show code"
    remove-header-space: true
    css: closeread-standalone.css
vignette: >
  %\VignetteIndexEntry{The Risk Landscape: A Scrollytelling Guide}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", echo = FALSE,
  fig.width = 8, fig.height = 5, dpi = 150,
  warning = FALSE, message = FALSE
)
if (requireNamespace("micromort", quietly = TRUE)) {
  library(micromort)
} else {
  pkgload::load_all(quiet = TRUE)
}
library(ggplot2)
library(dplyr)

dark_theme <- theme_minimal(base_size = 14) +
  theme(
    plot.background = element_rect(fill = "#1a1a2e", colour = NA),
    panel.background = element_rect(fill = "#1a1a2e", colour = NA),
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.major.x = element_line(colour = "#333355"),
    text = element_text(colour = "#e0e0e0"),
    axis.text = element_text(colour = "#e0e0e0"),
    plot.caption = element_text(colour = "#888888", size = 10),
    legend.background = element_rect(fill = "#1a1a2e"),
    legend.text = element_text(colour = "#e0e0e0"),
    legend.title = element_text(colour = "#e0e0e0")
  )
```

:::{.cr-section}

### How dangerous is your morning?

Every morning you make dozens of tiny gambles with your life. Brewing coffee. Crossing the road. Driving to work.

Each carries a **micromort** — a one-in-a-million chance of death.[@cr-mundane]{focus-on="cr-mundane"}

Your cup of coffee? **0.01 micromorts.** The ride to work by car? **0.13 micromorts.** By bicycle? **0.5 micromorts** — roughly four times riskier than driving.

These numbers are vanishingly small. But they are not zero — and they are not equal.

:::{#cr-mundane}
```{r mundane-plot}
cr <- common_risks()
mundane <- cr |>
  filter(activity %in% c(
    "Cup of coffee", "Crossing a road",
    "Commuting by car (30 min)", "Commuting by bicycle (30 min)",
    "Taking a bath", "Working in an office (8 hours)",
    "Drinking a glass of wine")) |>
  arrange(micromorts) |>
  mutate(activity = factor(activity, levels = activity))

ggplot(mundane, aes(x = micromorts, y = activity)) +
  geom_col(fill = "#2c7be5", alpha = 0.9) +
  geom_text(aes(label = sprintf("%.2f mm", micromorts)),
            hjust = -0.1, size = 4, colour = "#e0e0e0") +
  scale_x_continuous(expand = expansion(mult = c(0, 0.3))) +
  labs(x = "Micromorts", y = NULL,
       caption = "1 micromort = one-in-a-million chance of death") +
  dark_theme
```
:::

:::


:::{.cr-section}

### The full spectrum

Now zoom out.[@cr-spectrum]{focus-on="cr-spectrum"}

The micromort scale spans **six orders of magnitude** — from 0.01 (a dental X-ray) to nearly 40,000 (summiting Everest).

At the bottom: the invisible risks of daily life. A cup of coffee, a bath, a road crossing. Each below 0.1 micromorts.

In the middle: the risks we weigh consciously. A hospital stay (75 mm). Surgery under general anaesthesia (10 mm). A long-haul flight (4 mm).

At the top: risks that make headlines. Base jumping (430 mm per jump). **Everest: 37,932 micromorts per ascent** — a 1-in-26 chance of death.

The logarithmic scale is not an aesthetic choice. It is the only honest way to show data that ranges from coffee to Everest.

:::{#cr-spectrum}
```{r spectrum-plot}
cr_top <- cr |>
  filter(micromorts > 0) |>
  arrange(desc(micromorts)) |>
  slice_head(n = 30) |>
  mutate(activity = factor(activity, levels = rev(activity)))

ggplot(cr_top, aes(x = micromorts, y = activity)) +
  geom_point(aes(colour = category), size = 3) +
  scale_x_log10(labels = scales::label_comma()) +
  scale_colour_viridis_d(option = "turbo") +
  labs(x = "Micromorts (log scale)", y = NULL, colour = "Category",
       caption = "Source: micromorts.rip, Wikipedia, CDC MMWR") +
  dark_theme +
  theme(axis.text.y = element_text(size = 10))
```
:::

:::


:::{.cr-section}

### The slow burn

Not all risks kill you in an instant.[@cr-chronic]{focus-on="cr-chronic"}

Some shorten your life by half an hour, every single day. David Spiegelhalter called these **microlives** — each one worth 30 minutes of life expectancy.

**Smoking 20 cigarettes a day costs 10 microlives.** That is 5 hours of life, lost every 24 hours. A heavy smoker lives as though the day were only 19 hours long.

But the scale works in reverse too. **150 minutes of weekly exercise gains 3 microlives per day.** Five servings of fruit and vegetables: +4 microlives. A Mediterranean diet: +2.

The slow burn is not all loss. Some of it is gain.

:::{#cr-chronic}
```{r chronic-plot}
ch <- chronic_risks()
ch <- ch |>
  arrange(microlives_per_day) |>
  mutate(
    factor = factor(factor, levels = factor),
    fill_colour = ifelse(direction == "gain", "#198754", "#dc3545")
  )

ggplot(ch, aes(x = microlives_per_day, y = factor)) +
  geom_col(aes(fill = fill_colour), alpha = 0.9) +
  scale_fill_identity() +
  geom_vline(xintercept = 0, colour = "#e0e0e0", linewidth = 0.5) +
  labs(x = "Microlives per day", y = NULL,
       caption = "1 microlife = 30 minutes of life expectancy. Red = lose, Green = gain.") +
  dark_theme +
  theme(axis.text.y = element_text(size = 9))
```
:::

:::


:::{.cr-section}

### The bridge

How do you compare a skydive to a cigarette? One kills instantly (rarely). The other kills slowly (certainly, at scale).[@cr-bridge]{focus-on="cr-bridge"}

The answer is **Loss of Life Expectancy** — the average time lost or gained per exposure.

One micromort costs roughly **21 minutes** of life expectancy. One microlife is **30 minutes**. So 1 micromort ≈ 0.7 microlives.

The UK Department for Transport values a micromort at **£1.60**. NICE values a microlife at **£1.70**. Two independent government agencies, nearly the same price for a unit of risk.

:::{#cr-bridge}

| Unit | Definition | Example |
|------|-----------|---------|
| **Micromort** | 1-in-a-million death probability | Skydiving (per jump) = 10 mm |
| **Microlife** | 30 min of life expectancy per day | Smoking 2 cigs/day = −1 ml/day |
| **LLE** | Loss of Life Expectancy | 1 mm ≈ 21 min ≈ 0.7 microlives |
| **£ value** | NICE/DfT policy valuation | 1 mm ≈ £1.60 ≈ 1 ml |

: The three currencies of risk. {.striped}

:::

:::


:::{.cr-section}

### Test yourself

You now have three currencies for risk — micromorts, microlives, and loss of life expectancy.[@cr-quiz]{focus-on="cr-quiz"}

You know that a cup of coffee is 0.01 micromorts and Everest is 37,932. You know that smoking costs 10 microlives a day and exercise gains 3.

But can you tell which is riskier: a night in hospital or a skydive? Is red meat worse than sitting all day? Try the quizzes.

:::{#cr-quiz}

::: {style="background:#2b3e50;border-radius:12px;padding:24px;color:#fff;text-align:center;"}

### Can you rank these risks?

[**☠️ Micromort Quiz**](micromort-quiz.html){.btn .btn-danger .btn-lg style="margin:8px;"}
[**⏳ Microlife Quiz**](microlife-quiz.html){.btn .btn-success .btn-lg style="margin:8px;"}
[**📊 Rank Risks Quiz**](risk-ranking-quiz.html){.btn .btn-primary .btn-lg style="margin:8px;"}

:::

:::

:::

<details>
<summary>Session Info</summary>
```{r session-info, eval=TRUE}
sessionInfo()
```
</details>

