---
title: "Project Telemetry"
format:
  html:
    code-fold: true
    code-summary: "Show code"
vignette: >
  %\VignetteIndexEntry{Project Telemetry}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r telemetry-setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

if (requireNamespace("micromort", quietly = TRUE)) {
  library(micromort)
} else {
  pkgload::load_all(quiet = TRUE)
}

library(targets)
library(DT)

# Shared safe_tar_read with RDS fallback (see inst/vignette_utils.R)
source(file.path(tryCatch(rprojroot::find_root(rprojroot::is_r_package),
  error = function(e) "."), "inst", "vignette_utils.R"))
```

This vignette tracks the health and evolution of the **micromort** pipeline, codebase, and GitHub activity. All data is pre-computed via `targets` --- zero inline computation.

## 1. Pipeline Metrics

### Dependency Graph

```{r telemetry-dep-graph}
#| echo: false
#| results: asis
#| fig-cap: "Target dependency graph showing data flow across the micromort pipeline. Nodes represent individual targets; edges show upstream dependencies. The graph spans data acquisition, normalization, export, logging, documentation, telemetry, and vignette output stages. Built from `targets::tar_network()` edges. See the introduction vignette for a conceptual overview of the pipeline architecture."
#| fig-alt: "Directed graph of targets pipeline dependencies, with nodes representing targets and edges showing data flow from data acquisition through normalization, export, logging, documentation, telemetry, and vignette outputs."
show_target("vig_pipeline_dependency_graph")
```

### Plans and Targets

```{r telemetry-plans-targets}
#| echo: false
#| results: asis
show_target("vig_pipeline_summary")
```

### Top 5 Targets by Storage Size

```{r telemetry-top-by-size}
#| echo: false
#| results: asis
show_target("vig_telemetry_top_by_size")
```

### Top 5 Targets by Compute Time

```{r telemetry-top-by-time}
#| echo: false
#| results: asis
show_target("vig_telemetry_top_by_time")
```

## 2. GitHub Activity

### Commit Velocity

```{r telemetry-commit-velocity}
#| echo: false
#| results: asis
#| fig-cap: "Weekly commit frequency over the last 26 weeks. Each bar represents commits merged in a calendar week. Data sourced from `gert::git_log()` with a 500-commit lookback window. Weeks with zero commits indicate maintenance pauses or release stabilization periods. Compare with GitHub Activity table below for issue/PR context."
#| fig-alt: "Bar chart showing number of commits per week over the last 6 months, with weeks on the x-axis and commit count on the y-axis."
show_target("vig_telemetry_commit_velocity_chart")
```

### Issues and Pull Requests

```{r telemetry-github-activity}
#| echo: false
#| results: asis
show_target("vig_github_activity")
```

## 3. Project Structure

```{r telemetry-codebase-metrics}
#| echo: false
#| results: asis
show_target("vig_codebase_metrics")
```

## Reproducibility

```{r telemetry-session-info, eval=TRUE}
sessionInfo()
```

```{r build-info}
#| echo: false
#| results: asis
cat(safe_tar_read("vig_build_info") %||% "")
```
