---
title: "Project Telemetry"
format:
  html:
    code-fold: true
    code-summary: "Show code"
vignette: >
  %\VignetteIndexEntry{Project Telemetry}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r telemetry-setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

if (requireNamespace("micromort", quietly = TRUE)) {
  library(micromort)
} else {
  pkgload::load_all(quiet = TRUE)
}

library(targets)
library(DT)

# Safe tar_read with graceful fallback (finds project root for pkgdown builds)
safe_tar_read <- function(name) {
  store <- Sys.getenv("TARGETS_STORE", unset = NA)
  if (is.na(store)) {
    store <- tryCatch(
      file.path(rprojroot::find_root(rprojroot::is_r_package), "_targets"),
      error = function(e) "_targets"
    )
  }
  tryCatch(
    targets::tar_read_raw(name, store = store),
    error = function(e) {
      message("Target '", name, "' not found. Run tar_make() first.")
      NULL
    }
  )
}
```

This vignette tracks the health and evolution of the **micromort** pipeline, codebase, and GitHub activity. All data is pre-computed via `targets` --- zero inline computation.

## 1. Pipeline Metrics

### Dependency Graph

```{r telemetry-dep-graph}
#| fig-cap: "Target dependency graph showing data flow across the micromort pipeline. Nodes represent individual targets; edges show upstream dependencies. The graph spans data acquisition, normalization, export, logging, documentation, telemetry, and vignette output stages. Built from `targets::tar_network()` edges. See the introduction vignette for a conceptual overview of the pipeline architecture."
#| fig-alt: "Directed graph of targets pipeline dependencies, with nodes representing targets and edges showing data flow from data acquisition through normalization, export, logging, documentation, telemetry, and vignette outputs."
dep_graph <- safe_tar_read("vig_pipeline_dependency_graph")
if (!is.null(dep_graph)) {
  # Render mermaid via htmltools
  htmltools::HTML(paste0(
    '<script src="https://cdn.jsdelivr.net/npm/mermaid@11/dist/mermaid.min.js"></script>',
    '<script>mermaid.initialize({startOnLoad: true, securityLevel: "loose", theme: "dark"});</script>',
    '<div class="mermaid" style="background-color: #000000; padding: 1em;">\n', dep_graph, '\n</div>'
  ))
} else {
  htmltools::p("Dependency graph not available. Run tar_make() first.")
}
```

### Plans and Targets

```{r telemetry-plans-targets}
summary <- safe_tar_read("vig_pipeline_summary")
if (!is.null(summary)) {
  summary$plans |>
    DT::datatable(
      caption = paste0(
        "Target counts by plan file (",
        summary$total_targets, " total targets)"
      ),
      options = list(pageLength = 10, dom = "t"),
      rownames = FALSE
    )
}
```

### Top 5 Targets by Storage Size

```{r telemetry-top-by-size}
if (!is.null(summary) && !is.null(summary$top_by_size)) {
  summary$top_by_size |>
    dplyr::select(name, size_kb) |>
    DT::datatable(
      caption = "Largest targets by stored object size (KB)",
      options = list(pageLength = 5, dom = "t"),
      rownames = FALSE
    )
}
```

### Top 5 Targets by Compute Time

```{r telemetry-top-by-time}
if (!is.null(summary) && !is.null(summary$top_by_time)) {
  summary$top_by_time |>
    dplyr::select(name, seconds) |>
    DT::datatable(
      caption = "Slowest targets by computation time (seconds)",
      options = list(pageLength = 5, dom = "t"),
      rownames = FALSE
    )
}
```

## 2. GitHub Activity

### Commit Velocity

```{r telemetry-commit-velocity, fig.width=10, fig.height=5}
#| fig-cap: "Weekly commit frequency over the last 26 weeks. Each bar represents commits merged in a calendar week. Data sourced from `gert::git_log()` with a 500-commit lookback window. Weeks with zero commits indicate maintenance pauses or release stabilization periods. Compare with GitHub Activity table below for issue/PR context."
#| fig-alt: "Bar chart showing number of commits per week over the last 6 months, with weeks on the x-axis and commit count on the y-axis."
velocity <- safe_tar_read("vig_commit_velocity")
if (!is.null(velocity)) {
  ggplot2::ggplot(velocity, ggplot2::aes(x = week, y = commits)) +
    ggplot2::geom_col(fill = "#1976D2") +
    ggplot2::labs(
      title = "Commit Velocity",
      subtitle = "Last 26 weeks",
      x = "Week",
      y = "Commits"
    ) +
    ggplot2::theme_minimal() +
    ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 45, hjust = 1))
}
```

### Issues and Pull Requests

```{r telemetry-github-activity}
activity <- safe_tar_read("vig_github_activity")
if (!is.null(activity)) {
  activity |>
    DT::datatable(
      caption = "GitHub issues and pull requests summary",
      options = list(pageLength = 5, dom = "t"),
      rownames = FALSE
    )
}
```

## 3. Project Structure

```{r telemetry-codebase-metrics}
metrics <- safe_tar_read("vig_codebase_metrics")
if (!is.null(metrics)) {
  metrics |>
    DT::datatable(
      caption = "Codebase metrics snapshot",
      options = list(pageLength = 10, dom = "t"),
      rownames = FALSE
    )
}
```

## Reproducibility

```{r telemetry-session-info, eval=TRUE}
sessionInfo()
```
